\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\input{../common/preamble}
\input{../common/macros}

% Draft build: `make draft` drops draft.flag to surface TODO/CITE/NOTE markers.
% Submission builds leave it absent, so the markers vanish.
\IfFileExists{draft.flag}{\draftmodetrue}{}

\title{RCU Pseudo-Transactions and the Single-Writer Flip-Latch:\\
       Atomic Multi-Slot Publication for Read-Copy-Update Data Structures}

\author{Mathieu Desnoyers\\ EfficiOS Inc.\\ \texttt{mathieu.desnoyers@efficios.com}}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
% ---------------------------------------------------------------------------
% FRAMING RULE FOR THE WHOLE SERIES -- do not soften this in editing.
% The guarantee is: atomic multi-slot write, opt-in per-slot read validation.
% NEVER "serializable", NEVER "opacity", NEVER "STM". "Linearizable" stays
% scoped to the commit itself, never to a read+write transaction. Overclaiming
% here is the single most damaging error available to this paper.
% ---------------------------------------------------------------------------
Read-copy-update (RCU) makes a \emph{single} pointer store atomic with respect
to concurrent readers, and readers pay nothing for it. This works only when an
update is expressible as one store. Updates that must change several
words coherently---a doubly-linked list's back edge, a skip list's tower---fall
back to copying a larger region, a grace-period drain, or omitting the
operation. Linux's list does the last: it ships no RCU-safe reverse iterator,
because \code{\_\_list\_del()} publishes its forward edge with a marked store
and its back edge with a plain assignment. Userspace RCU's is identical.

We describe an RCU \pseudotxn: a frozen set of \record{slot}{old}{new} records
made visible by one commit step, together with a \sw realization---the
\fliplatch---in which every proxy in a group names one shared selector word, so
a single store switches the whole set. Two transacted structures follow: a
bidirectional list whose two chains stay mutually coherent, making the reverse
walk ordinary, and a hash list.

The guarantee is deliberately weaker than software transactional memory. The
write set is atomic: every record resolves to its old value before the commit
and to its new value after. But a reader is \emph{not} a transaction of any
kind---each of its loads resolves independently, so a traversal spanning several slots may
straddle a commit. There is no opacity and no multi-read snapshot; a reader gets
monotonicity, never new-then-old.

The weakening is the design. Where a structure is mutated in place, isolation is
a read-side cost: state on every element, work on every dereference---and the
read side is what RCU exists to keep free. A \pseudotxn's marker instead rides
in spare bits of the slot itself, the word the reader loads anyway, so it adds
no per-element state and no extra load, and vanishes when no mutation is in
flight. The cost moves to the writer, in the width of a commit.
\end{abstract}

% ===========================================================================
\section{Introduction}
\label{sec:intro}

Read-copy-update rests on a bargain. Readers pay nothing --- no lock, no
barrier, no write to shared state --- and in exchange a writer gets exactly one
atomic act: a single pointer store, bracketed by building the new structure
where readers cannot reach it and reclaiming the old one after a grace period.
The bargain has held for decades because the structures \rcu serves are
read-dominated, so making the majority free is worth constraining the minority.

The constraint is real, though, and it is one store. An update that must change
several words coherently has no atomic act to express itself with. The remedies
are familiar and each gives something up: copy the enclosing region so the
change becomes one pointer swap, and pay to copy; drain the readers with a grace
period so nobody spans the inconsistent window, and pay to wait; or decline the
operation.

\Cref{sec:rculistgap} looks at a library that took the third option, and it is
the cleanest evidence available that the problem is real, and it is not confined
to a niche library. Linux's \rcu list --- the most widely deployed there is ---
publishes only its forward chain: \code{\_\_list\_del()} commits the forward
edge with \code{WRITE\_ONCE()} and writes the back edge with a plain
assignment, and \code{\_\_list\_add\_rcu()} writes it \emph{after} the release
that publishes the node. Userspace \rcu's list has the identical shape. The
consequence is not a subtle ordering nit but a semantic one: the two chains are
changed by two stores and nothing makes them one event, so there is an interval
in which they describe different lists --- and a removed node survives as a
ghost until its grace period elapses, so the \code{prev} view can trail the
\code{next} view by that long. No barrier fixes either problem. So both
libraries ship forward iterators and no \rcu-safe reverse walk, and Linux goes
further, poisoning the back pointer in \code{list\_del\_rcu()} so it cannot be
followed at all. Nobody has to be persuaded the gap matters: the missing API is
the argument.

\subsection*{Contributions}

This paper describes an engine that gives \rcu writers a second atomic act ---
a set of word-sized slots that change together --- without disturbing the
bargain that makes \rcu worth using. Specifically:

\begin{itemize}
\item \textbf{A \pseudotxn model} (\cref{sec:model}): a frozen set of
      \record{slot}{old}{new} records made visible by one commit step. We state
      what it guarantees, and state what it does not before any evaluation ---
      including the property that a reader is not a transaction of any kind, and
      the weaker
      one that it does have: monotone resolution, never new-then-old
      (\cref{sec:monotone}).

\item \textbf{The \fliplatch} (\cref{sec:fliplatch}): a \sw realization in which
      every proxy in a group names one shared selector word, so a single release
      store switches the set. It has no compare-and-swap, no conflict detection
      and no abort --- and correspondingly requires writer exclusion, which we
      state as a precondition of the mechanism rather than a caveat on it.

\item \textbf{A per-record tag scheme} (\cref{sec:tags}): the marker lives in
      spare bits of the slot rather than in per-element state, which is what
      takes the reader's added loads to zero --- the test is on a register the
      reader already holds, and costs it only a predicted branch. Because a
      \pseudotxn may span structures
      whose encodings differ, the tag must travel with the record rather than be
      fixed per translation unit. Slots need not hold pointers: a shifted
      integer or a bitmap word is transacted the same way.

\item \textbf{Two transacted structures} (\cref{sec:wrappers}): a bidirectional
      list whose two chains stay mutually coherent --- so the reverse walk
      \code{rculist} declines to offer becomes ordinary --- and a hash list.
      Several edits compose into one commit, including edits that touch the same
      slots, through a read-your-own-writes discipline (\cref{sec:swcomposition});
      we give the two obligations that remain.

\item \textbf{The design space we did not take} (\cref{sec:variations}):
      selector encodings, tag placement and width, proxy allocation, settle
      policies, reclamation schemes and non-pointer encodings, with the cost of
      each alternative rather than only the name of it.
\end{itemize}

\paragraph{Scope.} This paper covers the \sw case, where writers are mutually
excluded by the embedder. A companion paper covers the concurrent engine, which
attacks the same model with a multi-word compare-and-swap and pays for
contention detection with machinery this one does not need. The two share the
model of \cref{sec:model} and the status vocabulary, deliberately: an embedder
written against one migrates to the other without rewriting its commit handling.
Properties that hold only under exclusion are marked where they appear.

\subsection{Where the cost goes}
\label{sec:costcomparison}

Multi-pointer atomicity under \rcu is not a new goal, and the established
answers pay for it in the same place: state attached to every element, and work
on the reader's path. \Cref{tab:comparison} places this engine against them.

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}l l c c l l@{}}
\toprule
 & \textbf{Per-element} & \textbf{Extra} & \textbf{Extra} & \textbf{Reader} & \textbf{Publish} \\
 & \textbf{state} & \textbf{loads} & \textbf{branches} & \textbf{guarantee} & \textbf{step} \\
\midrule
Existence~\citep{mckenney2016beyondissaquah}
  & \code{eh\_egi} field
  & 1
  & 1
  & per-element
  & group state \\
RLU~\citep{matveev2015rlu}
  & object header
  & 1
  & 1
  & snapshot
  & clock bump \\
MV-RLU~\citep{kim2019mvrlu}
  & header $+$ versions
  & $\geq 2$
  & $\geq 3$
  & snapshot
  & clock bump \\
\midrule
\textbf{This work}
  & \textbf{none}
  & \textbf{0}
  & 1
  & never new-then-old
  & one store \\
\bottomrule
\end{tabular}
\caption{Where each design pays for multi-pointer atomicity, on the read-side
fast path, per dereference. \emph{Extra loads} counts additional memory
references over and above the load the reader must issue anyway; \emph{extra
branches} counts conditional branches. Every design tests something, this one
included --- the honest difference is in the loads column, not the branches
column. Existence's load is an \emph{acquire} of a per-element field; RLU's is a
plain load from its per-thread state. A \pseudotxn's is absent because its
marker rides in the value \code{v} that \code{rcu\_dereference} already
returned, so its test is arithmetic on a register the reader already holds ---
but the test is still a branch, and \textbf{0} in the loads column means zero
loads, not zero instructions. All three single-branch entries are predicted
branches, statically hinted in their respective sources, and each costs a
branch-predictor entry. The first column is the unconditional cost, and is where
the designs differ permanently: existence and RLU attach state to every element
for as long as the structure lives, whereas a \pseudotxn attaches nothing and
confines its state to the mutation. MV-RLU is the one entry whose counts are not
constants: its dereference reads the object header unconditionally --- it has no
counterpart to RLU's pure-reader bypass --- and then may walk a version chain
whose length the reader does not control, so its figures are floors rather than
typical cases.}
\label{tab:comparison}
\end{table}

% ---------------------------------------------------------------------------
% PROVENANCE OF EVERY CELL IN tab:comparison -- VERIFIED FROM SOURCE 2026-07-16.
% Kept as a LaTeX comment rather than a \NOTE because it quotes C: the bare _, &
% and ~ would need escaping inside a macro argument, and an escaping slip here
% breaks only the draft build, which is exactly when you need to read this.
% Do not "tidy" these cells from memory. What was actually read:
%
% existence_exists()  --  perfbook datastruct/existence/existence.h:116
%     uintptr_t egi = smp_load_acquire(&ehp->eh_egi);
%     if (likely(!egi)) return 1;
%     esp = (uintptr_t *)(egi & ~0x1);
%     return (egi & 0x1) == smp_load_acquire(esp);
% The FAST path is ALREADY one acquire load of a PER-ELEMENT field (eh_egi).
% When the element is in a group there is a SECOND acquire load, of the group's
% state word -- and struct existence_group is __attribute__((aligned(
% CACHE_LINE_SIZE))), so that is a line shared by every reader of every element
% in the group. Note also: existence uses pointer tagging in bit 0 of eh_egi
% ("Dmitry Vjukov's pointer-tagging trick", per its own comment). Say so.
% Tagging is not our invention and claiming it would be trivially refuted.
%
% RLU_DEREF_INTERNAL  --  third_party/rlu/rlu.h:221
%     if (rlu_likely(self->is_check_locks == 0)) { p_cur_obj = p_obj; }
%     else { if (p_obj && RLU_IS_UNLOCKED(p_obj)) ... else slow_path }
% and rlu.c:797 clears is_check_locks at reader-lock time when the thread has no
% pending writes of its own (n_pure_readers++). So a PURE READER never touches
% the object header: a thread-local load and a predicted branch, nothing more.
%
% DO NOT claim "RLU costs an extra load on every read". It is FALSE for pure
% readers and a reviewer who knows RLU will catch it on sight. RLU's
% unconditional cost is SPACE -- rlu_obj_header_t is prepended to every object
% (RLU_MOVE_PTR_BACK) -- not instructions. An earlier draft of this table made
% exactly that error; it was caught only by reading rlu.h.
%
% mvrlu_deref()  --  github.com/cosmoss-jigu/mv-rlu, lib/mvrlu.c, read 2026-07-16
%     p_act = get_act_obj(obj);                    /* reads the object header */
%     p_copy = vobj_to_obj_hdr(p_act)->p_copy;     /* second header load      */
%     if (unlikely(p_copy)) {
%         do {
%             chs = vobj_to_chs(p_copy);
%             wrt_clk = get_wrt_clk(chs);
%             if (lte_clock(wrt_clk, self->local_clk)) return p_copy;
%             if (unlikely(lte_clock(chs->cpy_hdr.wrt_clk_next, qp_clk2)))
%                     break;
%             p_copy = chs->obj_hdr.p_copy;        /* version-chain walk      */
%         } while (p_copy);
%     }
%     return p_act;
%
% So: >= 2 loads, and NOT a constant. Two facts matter and both were verified,
% not inferred from the RLU row (the comment above says why that inference is
% unsafe -- and it would have been WRONG here in the opposite direction):
%
%   1. The header read is UNCONDITIONAL. There is no counterpart to RLU's
%      `if (self->is_check_locks == 0)` pure-reader bypass. Multi-versioning
%      costs RLU's pure-reader fast path -- MV-RLU's reader always touches the
%      object. This is a real difference BETWEEN the two RLU rows.
%   2. The version-chain walk is length-dependent with no iteration limit, so
%      the cell is a floor, not a typical case. That is why it reads ">= 2"
%      rather than a number.
%
% BRANCH COLUMN -- from the same sources quoted above, all fast paths:
%   existence  if (likely(!egi)) return 1;                     -> 1, hinted
%   RLU        if (rlu_likely(self->is_check_locks == 0))      -> 1, hinted
%   MV-RLU     if (unlikely(!obj)) / is_obj_actual() inside
%              get_act_obj() / if (unlikely(p_copy))           -> >= 3, + loop
%   this work  urcu_txn_sw_list_resolve(), rcu-txn-sw-list.h:
%                 uintptr_t v = (uintptr_t) ptr;
%                 if (caa_unlikely(v & URCU_TXN_SW_LIST_PROXY_TAG)) {...}
%                 return ptr;                                  -> 1, hinted
%
% The point of carrying this column at all: EVERY design branches, ours
% included, so the branch is NOT where the comparison is decided and claiming
% otherwise would be unfair in our favour. The difference is that our test's
% operand is already in a register while everyone else must load theirs first.
% Being scrupulous here strengthens the argument rather than weakening it --
% do not drop the column to make the row look cleaner. Note also that our 1 is
% deliberately NOT bolded: bold marks where this design wins, and it does not
% win here.
%
% Deliberately NOT claimed: mvrlu_deref also does self->num_act_obj++ and
% self->num_deref++ (per-thread STORES on the read path). Those read like
% statistics counters and may be compiled out in a production build; we did not
% verify that, so we do not count them. Counting them would be exactly the kind
% of flattering-to-us overclaim the RLU note above exists to prevent.
%
% RLU HAS NO TLS. There is no __thread anywhere in rlu.h/rlu.c: `self` is a
% caller-held pointer parameter, so self->is_check_locks is a plain load through
% a register. Do not write "thread-local flag" -- it is a field of a per-thread
% STRUCT, which is not the same claim.
%
% Our own benchmark is what puts `self` in TLS (bench_list_scale.c:1505,
% `static __thread rlu_thread_data_t *tls_rlu`), fetched once per operation via
% rlu_self() at :1579 and then held in a register across the traversal. So it is
% hoisted out of the per-dereference path this table is about.
%
% DO NOT argue TLS access models (initial-exec vs global-dynamic under dlopen)
% against RLU. It is a property of how an embedder stores `self`, not of RLU's
% design -- RLU takes `self` as a parameter precisely to leave that open -- and
% it is amortised per operation, not per dereference. It also cuts both ways:
% rcu_read_lock() touches URCU_TLS(rcu_reader) under the same rules (urcu ships
% tls-compat.h for exactly this), so the point nets to ~zero and reads as
% reaching. The argument is strongest on ground where the difference is
% structural: existence and RLU must put their mechanism somewhere the reader
% can LOAD it; ours is already in the register.
% ---------------------------------------------------------------------------

% ===========================================================================
\section{Background}
\label{sec:background}

This section fixes only the vocabulary the rest of the paper needs. Readers
familiar with \rcu will find nothing new in \cref{sec:rcu} and
\cref{sec:buildpublish}; \cref{sec:onestore} is where the problem is stated, and
it is stated on shipped code rather than in the abstract.

\subsection{Read-copy-update}
\label{sec:rcu}

\rcu~\citep{mckenney1998rcu,desnoyers2012urcu} divides a mutation between two
parties and gives each of them a different bargain. A reader delimits its
traversal with \code{rcu\_read\_lock()} and \code{rcu\_read\_unlock()}, marking an
\rcu \emph{read-side critical section}. On the
common implementations these are not locks at all: they exclude nothing and
contend with nothing, and in the userspace library's \QSBR mode
\code{rcu\_read\_lock()} is an empty function --- a debug assertion and nothing
else. What a reader gets in exchange is narrow but sufficient: any object it
reaches inside the critical section remains allocated until that section ends.

That is not magic, and it is worth saying where the cost went rather than
leaving an empty function to look like a free lunch. \QSBR buys its empty read
lock by moving the reader's obligation out of the critical section and onto the
thread: a
registered thread must periodically call \code{rcu\_quiescent\_state()} to
announce that it holds no \rcu-protected references, and must call
\code{rcu\_thread\_offline()} before blocking for any length of time. The
announcement is cheap --- it compares the global grace-period counter against
the thread's own and returns immediately when they agree --- but somebody has to
make it, and the writer's grace period cannot end until every registered thread
has. So the cost is not eliminated; it is amortized across a whole thread's
worth of read-side critical sections and relocated to a point the application
chooses.
The price is structural rather than instruction-level: the application must
\emph{have} such points and must be written to reach them promptly, which is why
\QSBR is at once the fastest flavour and the most intrusive to adopt. The other
flavours make the opposite trade, paying a little inside the critical section so
the application owes nothing between them.

That promise is kept by the writer, not by the reader, and it is kept by
waiting. A \emph{grace period} is an interval that every pre-existing read-side
critical section has been observed to leave. Once one elapses, no reader can still hold a
reference obtained before it began, so anything unlinked before it began is
unreachable and may be freed. The writer waits synchronously
(\code{synchronize\_rcu()}) or defers a callback (\code{call\_rcu()}); either
way the cost of reclamation is charged entirely to the side doing the mutating.

Publication is the mirror image. A writer makes an object reachable with
\code{rcu\_assign\_pointer()}, a release store, and a reader reaches it with
\code{rcu\_dereference()}, which returns the pointer and guarantees that the
loads dependent on it see the object's initialized contents rather than
pre-initialization garbage. On every architecture Userspace \rcu supports, the
reader's half of that is free: the dependency between loading a pointer and
loading through it is respected by the hardware, so \code{rcu\_dereference()}
emits an ordinary load and no barrier.\footnote{Alpha, which did not respect it,
is the standing exception and is no longer supported. The C11 \code{consume}
ordering that was meant to express this in the abstract machine is not usable in
practice, and compilers promote it to \code{acquire}~\citep{boehm2008memorymodel,batty2011cppconcurrency};
Userspace \rcu therefore relies on the dependency directly.}

The whole arrangement rests on one sentence: \emph{readers pay nothing, and in
exchange a writer gets exactly one atomic act.} The rest of this paper is about
the second half of that sentence.

\subsection{Build, publish, reclaim}
\label{sec:buildpublish}

The one atomic act is a single pointer store, and it sits in the middle of a
three-phase discipline that every \rcu mutation follows:

\begin{enumerate}
\item \textbf{Build.} Allocate and fully initialize the new nodes \emph{where no
reader can reach them}. Because they are unreachable, this phase is
single-threaded by construction: no ordering is needed, no barrier is paid, and
an operation that fails here has published nothing and can free its work
immediately with no grace period.
\item \textbf{Publish.} One release store makes the new structure reachable.
This is the linearization point, and it is the only instant at which concurrent
readers change what they see. It is also the paper's pivot: \emph{everything
before it is private, everything after it is committed, and there is exactly one
of it.}
\item \textbf{Reclaim.} The nodes the publish displaced were reachable, so they
may still be held by readers that started earlier. Free them after a grace
period, never before.
\end{enumerate}

The middle phase is the constraint. Build is unconstrained and reclaim is merely
deferred, but publish must be \emph{one store}, because a store is the only
thing the hardware will make atomic with respect to a reader that has taken no
lock and paid no barrier.

This frame carries through the whole series, and it is worth stating the
punchline now so that \cref{sec:model} lands as an extension rather than a
replacement: a \pseudotxn does not add a fourth phase. It \emph{widens the
second one}. Build and reclaim keep their meanings exactly; commit is the
publish step, still a single store, and what changes is how many slots that one
store moves.

\subsection{Why one store is not enough}
\label{sec:onestore}

An update that must re-point two edges coherently has no atomic act to express
itself with, and the consequence is not hypothetical. It is visible in the
doubly-linked list of the most widely deployed \rcu implementation there is: the
one in the Linux kernel~\citep{linux2026rculist}.

Removing a node \code{E} from between \code{P} and \code{N} re-points two
reader-visible edges, and the kernel writes each with a different kind of store:

\begin{lstlisting}
static inline void __list_del(struct list_head *prev, struct list_head *next)
{
        next->prev = prev;              /* N->prev: E -> P   plain store  */
        WRITE_ONCE(prev->next, next);   /* P->next: E -> N   marked store */
}
\end{lstlisting}

The forward edge gets \code{WRITE\_ONCE()}. The back edge gets a bare \code{=},
to a node readers can reach at that very moment. Userspace \rcu's
\code{cds\_list\_del\_rcu()} has the identical shape, down to the ordering of
the two lines. And the asymmetry is not an accident of the delete path ---
\code{\_\_list\_add\_rcu()} makes it starker still:

\begin{lstlisting}
new->next = next;                              /* private: unreachable */
new->prev = prev;                              /* private: unreachable */
rcu_assign_pointer(list_next_rcu(prev), new);  /* release: publishes new */
next->prev = new;                              /* plain store, AFTER publish */
\end{lstlisting}

The forward edge is published with a release, and the back edge is written
\emph{afterwards}, with a plain store, to a node that is already reachable.
Between those two lines there is an interval in which \code{prev->next == new}
while \code{next->prev == prev}: the forward chain contains the new node and the
backward chain skips it. This needs no weak-memory argument --- it is plain
program order, so the window is real on \emph{every} architecture.

The kernel is candid about the consequence. \code{list\_del\_rcu()} does not
merely leave the back pointer unmaintained; it destroys it,
\code{entry->prev = LIST\_POISON2}, and the \rcu list ships forward iterators
only. Userspace \rcu draws the same line by omission: its plain list offers
\code{cds\_list\_for\_each\_prev()} and
\code{cds\_list\_for\_each\_entry\_reverse()}, and its \rcu list offers neither.
The backward chain exists in both libraries and every writer maintains it. A
reader may not walk it.

\paragraph{The kernel's narrow exception proves the rule.} In December 2024 the
kernel gained a way to read a \code{prev} pointer under
\rcu~\citep{brauner2024listbidir}: \code{list\_bidir\_del\_rcu()}, which skips
the poisoning, plus \code{list\_bidir\_prev\_rcu()}, an accessor. It was
suggested and reviewed by McKenney himself, so it is a considered answer rather
than an oversight being patched --- and it is worth being precise about what it
does and does not deliver, because it is the closest thing to a counterexample
this paper has.

It does not make the two chains agree. \code{\_\_list\_del()} still writes the
back edge with a plain store, and nothing orders that store against the forward
edge's \code{WRITE\_ONCE()}; \code{\_\_list\_add\_rcu()} still writes it after
the release. What the patch buys is that a deleted node keeps its own \code{prev}
instead of having it poisoned, so a reader standing on a ghost can step backward
out of it. That is a genuinely useful thing and it is not a coherent reverse
walk. Consistent with that, the kernel added no reverse iterator, and the API
costs a whole-list mode: \code{list\_del\_rcu()} and \code{list\_bidir\_del\_rcu()}
may not be used on the same list. Its sole in-tree caller as of
\code{next-20260226}, \code{\_\_ns\_tree\_adjoined\_rcu()} in
\code{kernel/nstree.c}, takes exactly one adjacency hop --- \code{prev}
\emph{or} \code{next} --- and never compares the two directions against each
other. It never needs coherence, so it never meets its absence.

\paragraph{Ordering is not what is missing.} It is tempting to read all of this
as an ordering bug: publish \code{prev} with \code{rcu\_assign\_pointer()}, read
it with \code{rcu\_dereference()}, and the reverse walk falls out. It does not,
and seeing why is the point of this subsection. Suppose both edges were properly
published. They are still \emph{two} release stores, hence two publication
instants, and a reader that arrives between them sees \code{P->next == N} while
\code{N->prev == E}. The forward chain has dropped \code{E}; the backward chain
still contains it. Neither view is corrupt and neither is torn. They simply
describe different lists, and a reader that walks forward and then hops backward
--- the operation a reverse iterator exists to serve --- gets an answer no
sequential list could ever produce.

What is missing is \emph{atomicity across two slots}, and \rcu has no way to ask
for it. That is why the fix has to be a mechanism rather than a barrier, and it
is why two libraries, independently, chose to withhold the operation instead.

There is a second, subtler failure that survives even careful ordering, and it
is specific to \rcu. An unlinked node is a ghost: it must remain allocated for a
grace period, and it keeps its own \code{next} and \code{prev} pointing at its
former neighbours. The live forward chain drops it at the instant of the
publish. A reverse walker stepping through \code{prev} pointers can keep landing
on it for as long as the ghost survives. The backward view therefore trails the
forward view by up to a grace period --- not by a few instructions, but by the
full reclamation latency of the algorithm. No amount of barrier placement closes
a window whose width is a grace period.

So the doubly-linked list, the textbook structure, is already past what one
store can express, and the library's response was to remove the operation from
the interface. \Cref{sec:fliplatch} makes both edges move at one instant, and
\cref{sec:wrappers} spends that to give the reverse walk back.

% ===========================================================================
\section{The Pseudo-Transaction Model}
\label{sec:model}

% ---------------------------------------------------------------------------
% SCOPE NOTE (drafting): this section is deliberately ENGINE-AGNOSTIC. It must
% define only what the single-writer flip-latch (Sec. 4) and the multi-writer
% MCAS engine (companion paper) BOTH realize, because the companion paper
% inherits these definitions verbatim. Anything that is a property of one
% realization -- descriptors, install/settle, the tri-state status word, abort
% -- belongs there, not here. In particular there is no abort in this model:
% under writer exclusion a commit cannot fail for contention, and the ABORT
% status exists only because the two front-ends share a status enum.
% ---------------------------------------------------------------------------

\subsection{Records and pseudo-transactions}

\begin{definition}[Record]
\label{def:record}
A \emph{record} is a triple \record{s}{o}{n}, where $s$ is the address of a
word-sized \emph{slot}, $o$ is the value the slot must hold at the commit
point, and $n$ is the value it takes afterwards.
\end{definition}

\begin{definition}[Pseudo-transaction]
\label{def:pseudotxn}
A \pseudotxn is a set of records over pairwise-distinct slots that is
\emph{frozen} --- closed to further records --- before a single \emph{commit
step} makes the whole set take effect at once.
\end{definition}

Two properties of \cref{def:pseudotxn} carry the entire model, and they concern
different things: \emph{what} takes effect, and \emph{when}. They also happen at
different times, and the order matters.

\emph{Freezing} settles the \emph{what}, and it comes first. The set is closed
before the commit step, not at it: once the commit procedure is under way the
membership is fixed, and no record may still join. So what will become visible
is decided in advance rather than accreting as the commit proceeds.

The \emph{single step} settles the \emph{when}, and it comes last. The commit
step is one store, not a protocol, so the already-frozen set takes effect all at
once --- there is no interval in which part of it is visible and part is not.

These are two distinct moments in sequence, and only the second is the commit
step. Staging the set is what closes it; the single store that follows is what
makes the closed set visible; and any work that tidies up afterwards changes
nothing a reader can see (\cref{prop:lp}). The staging and the tidying are not
part of the commit step and are not linearization points --- a fact easier to
see once the mechanism is concrete (\cref{sec:lifecycle}), where freezing
coincides with the start of installation and the commit step is the lone store
that flips the set from old to new.

\subsection{A transaction in the filesystem sense}
\label{sec:fssense}

The word \emph{transaction} carries two meanings, and only one of them applies
here. A journaling filesystem stages a batch of block writes, closes the batch,
and writes a commit record; before that record the old state is authoritative,
after it the new state is, and no partial batch is ever authoritative. It calls
this a transaction, and it has never meant isolation by it. A database or an
STM means something else: a unit that is atomic \emph{and} isolated, where
isolation constrains what concurrent \emph{readers} may observe.

Our model is the first of these, essentially point for point:

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Journaling filesystem} & \textbf{Pseudo-transaction} \\
\midrule
journal transaction, closed      & record set, frozen \\
journal blocks                   & \record{s}{o}{n} records \\
commit record                    & the commit step \\
before commit: old state         & before commit: every record resolves $o$ \\
after commit: new state          & after commit: every record resolves $n$ \\
never a partial batch            & no slot, and no instant, is ever torn \\
readers are not transactional    & readers are plain \rcu readers \\
gives atomicity, not isolation   & gives atomicity, not opacity \\
\bottomrule
\end{tabular}
\caption{The model is the journaling sense of \emph{transaction}, not the
database sense. The staged record set is the journal; the commit step is the
commit record.}
\label{tab:fssense}
\end{table}

This is why we say \pseudotxn rather than transaction. The prefix is not an
apology for an incomplete STM; it is a disambiguation for readers who take
\emph{transaction} to imply isolation. What the prefix marks as absent is
isolation, and the reason it is absent is the subject of
\cref{sec:whyweakening}: isolation is paid for on the read side, and the \rcu
read side is precisely what this engine declines to charge.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% If space is tight this table is the first thing to cut to prose --- but
% the point it makes is not optional. It is what stops a concurrency reader
% filing the model under "weak STM" in the first thirty seconds.
% -----------------------------------------------------------------------

\subsection{What is guaranteed}
\label{sec:guarantees}

\begin{property}[Write-set atomicity]
\label{prop:atomicity}
Every record in a committed \pseudotxn resolves to its old value $o$ before the
commit step and to its new value $n$ after it. One step switches the whole
frozen set.
\end{property}

\begin{property}[No tearing]
\label{prop:notearing}
No slot, and no instant, is ever torn: at every instant, every slot holds
either a value that resolves to $o$ or one that resolves to $n$, and never
anything else.
\end{property}

\begin{property}[The commit linearizes at one point]
\label{prop:lp}
The commit is a linearizable \emph{write operation}: it takes effect atomically
at a single instant, the commit step, which is its lone linearization
point~\citep{herlihy1990linearizability}. The staging that precedes it and the
tidying that follows change no observable value, so neither is a linearization
point.
\end{property}

The scope of \cref{prop:lp} is the write, and only the write --- and the
distinction is the reason for the \emph{pseudo-} prefix, not pedantry. A single-slot
read is itself a linearizable operation, taking effect at its own load. What is
\emph{not} a linearizable operation is a reader's traversal of several slots: it
is a sequence of loads, not one operation, so it has no linearization point of
its own and may straddle the commit (\cref{sec:nonguarantees}). Saying the write
linearizes is therefore a precise and limited claim. It is not the claim that a
combined read-and-write transaction is linearizable, and still less that it is
serializable --- neither holds, and neither is asserted anywhere in this paper.

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Keep "linearizable" scoped to the WRITE, exactly as in Property 3. The
% engine's own words: "This linearizes the WRITE." Never extend it to a
% read-plus-write bracket -- that bracket is not atomic and claiming
% otherwise is the single most damaging error available to this series.
% -----------------------------------------------------------------------

\begin{property}[Cross-structure atomicity]
\label{prop:crossstructure}
Because the record set may span data structures, the commit is one
linearization point \emph{across all of them}. This is strictly stronger than
composing independent \rcu structures, where a node can be reachable in $A$
before it is reachable in $B$. The unit of cross-structure atomicity is exactly
one \pseudotxn: two commits are two linearization points, the same as two
unrelated \rcu structures.
\end{property}

\paragraph{Worked example.} Consider a cache with two access paths: entries are
found by key through a hash index, and ordered by recency through a linked list.
Both are \rcu structures and both are read without a lock. Inserting an entry
$E$ has to make it findable and make it ordered.

Plain \rcu forces the writer to choose an order, and neither choice is right.
Publish into the index first, and there is an interval in which a reader finds
$E$ by key, follows it, and discovers it is in no position in the recency list.
Splice into the list first, and there is an interval in which $E$ occupies a
position in the recency list while a lookup for its key still reports it absent.
The interval is not a few instructions --- the writer can be preempted between
the two stores --- and no amount of barrier placement removes it, because there
genuinely are two publications. Callers are left to tolerate the skew, which
usually means the structures grow a validity flag, or a lock, and the reason
\rcu was chosen erodes.

A \pseudotxn names the slots of both structures in one record set: the index
bucket slot that must come to name $E$, and the list edges that splice it into
position. One commit publishes all of them. $E$ becomes findable and ordered at
the same instant, and the question of which to publish first stops existing,
because there is only one publish.

\Cref{prop:crossstructure} also fixes the boundary, and it is a real one. The
unit of cross-structure atomicity is one \pseudotxn, not one writer and not one
critical section. An operation that commits the index edit and then commits the
list edit has produced two linearization points and is back to exactly the skew
above --- the engine has made the composition \emph{available}, not automatic.
Folding edits into one commit is sound even when they touch the same slots: the
engine reconciles a writer's edits against its own pending ones through
read-your-own-writes (\cref{sec:swcomposition}). Cross-structure composition,
where the slots cannot coincide, is the easy case of that.

\subsection{What is not guaranteed}
\label{sec:nonguarantees}

We state the limits here, before any evaluation, because they are constitutive
of the model rather than defects in it.

% ---------------------------------------------------------------------------
% GUARDRAIL (do not delete in an editing pass).
%
% "a transaction of ANY KIND" -- never "a reader is not a \pseudotxn". This
% paper said the latter in five places and it is a genuine defect, not a style
% nit: "a reader is not a pseudo-transaction" parses just as naturally as "a
% reader is not MERELY a pseudo-transaction (it gets a real one)", which is the
% exact inverse of the claim and the most damaging misreading the paper admits.
%
% The drift is mechanical and WILL recur: the \pseudotxn macro is an inviting
% substitution for the bare word "transaction", and here the word must stay
% general precisely because the denial has to cover both classes at once. The
% \pseudotxn is a WRITER-side object; denying that a reader is one is nearly
% vacuous. The real claim is that the reader gets no transactional treatment
% whatsoever -- no bracket, no read set, no atomicity across loads.
%
% README's framing rule states it correctly; match it verbatim.
% ---------------------------------------------------------------------------
\textbf{A reader is not a transaction of any kind.} It is a sequence of
single-slot reads, each linearizing at its own access. It does have an \rcu
read-side critical section --- but a critical section is not a transaction. It
keeps reachable objects alive; it does not make the loads inside it a unit.
There is no read set and nothing that begins or commits on the read side: the
\pseudotxn is the writer's object, and a reader meets it one slot at a time. A
traversal spanning several slots may therefore \emph{straddle} a commit: it can
observe the \pseudotxn in a slot it reads after the commit and not in one it
read before.
\Cref{prop:notearing} still holds throughout --- no slot and no instant is torn
--- but a reader's several reads are simply not a mutual snapshot.

\textbf{There is no opacity and no multi-read snapshot.} This is weaker than
STM, deliberately and in exactly one direction: readers are plain \rcu readers
that pay no per-read barrier. An embedder that needs a multi-read snapshot
layers its own versioning on top, as with any value-based \mcas.

\textbf{A record's old is a value at an instant, not a value over an interval.}
\Cref{def:record}'s $o$ is what the slot must hold \emph{at} the commit point,
not what it must have held throughout. An operation whose correctness requires
\emph{detecting} an away-and-back change must carry its own version in the word.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Do NOT write "a validated slot is CHECKED to hold its expected value" in
% P1. The single-writer engine performs no check: it has no install CAS, so
% nothing verifies o against memory. Under exclusion the writer is entitled
% to know, and o is used for RESOLUTION (what a proxy returns pre-flip), not
% for validation. The concurrent engine's install CAS is what turns this
% into a check -- so the checking language belongs in P2, and the model here
% must state the property without implying an enforcement that does not
% exist. Sec. 7.4 explains where this DOES bite in P1: through composition.
% -----------------------------------------------------------------------

\subsection{Monotonicity: never new-then-old}
\label{sec:monotone}

The previous section understates what a reader gets, and the gap matters. A
reader's reads are not mutually atomic, but neither are they unrelated.

\begin{property}[Monotone resolution]
\label{prop:monotone}
The word that decides a \pseudotxn's outcome is written once and never written
back. A reader resolving several slots of one \pseudotxn over time therefore
observes a monotone sequence --- $o \ldots o, n \ldots n$ --- and
\textbf{never $n$ then $o$}. A reader may lag a commit; it never regresses
across one.
\end{property}

\Cref{prop:monotone} is strictly more than treating each slot as an independent
register and strictly less than a snapshot, and that middle ground is
exploitable rather than merely tolerable (\cref{sec:publishorder}).

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% DO NOT call this "causal". "Causal consistency" names a specific model
% that is WEAKER than linearizability -- it lets observers disagree on the
% order of unrelated writes -- so borrowing the word would make the engine
% read as claiming less than it provides. "Never new-then-old" is precise,
% memorable, collides with no existing model's name, and is the engine's own
% phrasing. Equally, do not write "per-slot linearizability": true of each
% read, but it implies reads are unrelated and so undersells Property 5.
% -----------------------------------------------------------------------

\paragraph{Scope.} \Cref{prop:monotone} has a precondition that must travel
with it. Monotonicity of the deciding word is unconditional: it is one word,
stored once. What a reader's \emph{sequence} of resolutions inherits from it
depends on how the reader got from one slot to the next. A
\emph{dependency-chained} traversal --- each hop's address derived from the
value the previous hop loaded --- carries the order for free on every
architecture \urcu supports, as does any build using the default C11
dereference. A reader that hops \emph{without} that chain, by re-reading a
pointer it cached earlier or by walking between two independently-reached
slots, has no such edge: under \code{-DURCU\_DEREFERENCE\_USE\_VOLATILE} on
weakly-ordered hardware its two loads may be reordered, and it can observe the
new value on the later slot and the old on the earlier one. Such a reader owes
itself an explicit acquire, or a dependency.

We state the build condition rather than relegating it, because it is the
difference between a guarantee and a hope. An embedder who misses it writes a
race that no test on a strongly-ordered machine will ever reproduce.

\subsection{Publish order, not snapshots}
\label{sec:publishorder}

\Cref{sec:nonguarantees} invites an obvious objection. If a reader gets no
snapshot, and a traversal touching several slots may straddle a commit, how is
any multi-slot traversal ever correct? The answer is not that snapshots are
recovered somewhere later. It is that a snapshot is only one way of buying what
the reader actually needs, and there is a cheaper way to buy it.

What a snapshot buys is \emph{independence from timing}: it makes a reader's
conclusions hold no matter when its individual loads happened. But a reader does
not read in an arbitrary order. It reads in the order its traversal dictates,
and that order is usually fixed by the structure --- a tree is descended, a list
is walked, a hash bucket is entered at its head. If the order in which edits
\emph{become visible} is made to agree with the order in which the reader
\emph{looks}, timing-independence stops being something to buy at all.

\paragraph{The rule.} Suppose a reader resolves slots in a fixed order
$s_1, s_2, \ldots, s_k$, and the writer splits the edit across several
\pseudotxns rather than one. Then:

\begin{quote}
\emph{Commit in the reverse of the reader's read order: publish $s_k$ first and
$s_1$ last.}
\end{quote}

The consequence is worth spelling out, because it is a short argument with a
disproportionate payoff. Write $t_i$ for the instant $s_i$ is published and
$a_i$ for the instant the reader resolves it. The reader's order gives
$a_1 < a_2 < \cdots < a_k$; the rule gives $t_1 > t_2 > \cdots > t_k$. Now
suppose the reader observes $s_i$ as \emph{new}, i.e.\ $a_i > t_i$. For any
later slot $s_j$ with $j > i$ we have $a_j > a_i > t_i > t_j$, so the reader
observes $s_j$ as new too. Once a reader crosses into the new state it stays
there for the remainder of its traversal. It sees a prefix of old and a suffix
of new, and the crossing point is wherever the commits caught it.

\paragraph{Why a prefix of old and a suffix of new is enough.} The two
observations the reader can be left with are not equally troublesome, and the
asymmetry is the whole point.

Old-then-new is a straddle, not a contradiction. It is precisely what an
ordinary \rcu reader has always tolerated: a reader that begins before a publish
sees the old structure, one that begins after sees the new, and one that
overlaps sees the old up to the point the publish caught it and the new
thereafter. Every deployed \rcu container already asks its callers to accept
this, and callers do, because the alternative --- excluding readers for the
duration of a traversal --- is the cost \rcu exists to avoid.

New-then-old is a different animal. A reader that observes the new value of one
slot and then the old value of another has seen two effects of a single state
transition in reverse order, and no sequential execution of the data structure
produces that. There is nothing the reader can do about it locally: it cannot
tell the stale load from a fresh one, so its only recourse is to re-read
everything and compare --- which is to say, to build a snapshot. \emph{This is
the observation that forces snapshots, and it is the one the model already
forbids.} \Cref{prop:monotone} rules it out by construction, because the
deciding word is written once and never written back.

So the design rule does not weaken the reader's obligations and then apologize
for it. It removes the only observation a reader genuinely could not have coped
with, and leaves behind exactly the straddle that \rcu programmers have been
handling correctly for two decades.

\paragraph{A worked example.} Re-parent a node $E$ from $P$ to $Q$ in a tree
that carries parent pointers. Two reader-visible edges move: the forward edge
($Q$'s child link now names $E$) and the back edge ($E$'s parent link now names
$Q$). A reader descends first and walks back up second, so it reads the forward
edge before the back edge. By the rule, publish them in the opposite order:

\begin{enumerate}
\item Commit the back edge: \code{E->parent}: \edit{P}{Q}.
\item Commit the forward edge: the child links of $P$ and $Q$.
\end{enumerate}

A reader that descends and finds $E$ in its new position under $Q$ read the
forward edge after the second commit, hence after the first, so when it walks
back up it finds \code{E->parent} already naming $Q$, and the two agree. The excluded
observation --- descending into the new position and then being told by the
parent pointer that $E$ still hangs off $P$ --- is exactly the
$(\mathit{new}, \mathit{old})$ pair the ordering makes unreachable. Publishing
the forward edge first would permit it.

\paragraph{Why this is safe, and where it stops.} The argument above quietly
assumed both orders it used, and both have to be earned. The writer's side is
free here: under exclusion (\cref{sec:exclusion}) the commits are two release
stores issued in program order by one thread. The reader's side is the one that
needs care, and it is exactly the precondition of \cref{sec:monotone} --- the
reader's loads must be genuinely ordered, not merely written in that sequence.

They are, and for a reason worth noticing rather than assuming: the reader
reaches $E$ by loading the forward edge, and then computes the address of
$E \rightarrow parent$ \emph{from the value that load returned}. The second load
depends on the first. The fixed-read-order idiom is therefore not something that
must be reconciled with the dependency-chaining requirement of
\cref{sec:monotone} --- it \emph{is} a dependency-chained descent, which is why
it is safe as stated and why it costs the reader no barrier.

The limits should be equally clear. This is an embedder's discipline, not an
engine guarantee: the engine cannot know a structure's traversal order, so
nothing checks that the commits were ordered against it. It applies only where a
read order actually is fixed; a reader that compares two independently-reached
slots, or re-reads a pointer it cached, has no chain to carry the order and is
back to owing itself an acquire. And it buys coherence along the traversal, not
a snapshot: the reader still cannot ask what the whole structure looked like at
one instant. What it can do --- and what a very large class of traversals needs
--- is never be told that time ran backwards.

That is what converts \cref{prop:monotone} from a consolation into a tool, and
it is the honest reply to the reviewer who asks why anyone would accept a model
without opacity.

\subsection{Why this weakening}
\label{sec:whyweakening}

\rcu exists because readers dominate. Its entire bargain is that a reader pays
nothing: no lock, no barrier, no write to shared state. Every design that adds
multi-pointer atomicity on top of \rcu must decide what to charge that reader,
and the answer is what separates them (\cref{tab:comparison}).

% ---------------------------------------------------------------------------
% GUARDRAIL (do not delete in an editing pass).
%
% The "only once you mutate in place" qualifier is load-bearing, not hedging.
% An earlier draft asserted flatly that "isolation is paid for on the read side"
% and argued it from "a snapshot is a claim about what a reader observes, so
% only the reader can establish it".
%
% That premise is FALSE, and this paper of all papers cannot afford it: Sec. 1
% already names copy-on-write as one of RCU's three remedies. A writer that
% path-copies and swaps a root gives readers a fully isolated immutable
% snapshot at ZERO read-side cost -- the reader establishes nothing. Persistent
% data structures are the existence proof, and RCU's own build/publish/reclaim
% is the same trick in the small. The unqualified claim contradicts our own
% Sec. 1 and a reviewer who knows persistent structures spots it immediately.
%
% The true claim is narrower and still sufficient: isolation is a read-side
% charge for designs that mutate IN PLACE, which is the regime every row of
% tab:comparison occupies, ours included. Do not "tighten" this back into the
% unqualified form -- it reads punchier and it is wrong.
% ---------------------------------------------------------------------------
\paragraph{Isolation is a read-side cost only once you mutate in place.} The
claim needs its precondition stated, because without it the claim is false and
this paper has already said why. There is a way to give readers a snapshot for
nothing, and \rcu uses it: copy. A writer that builds a new version where
readers cannot reach it and then swaps one pointer hands every reader an
immutable structure. The reader establishes nothing, validates nothing, and
still sees a coherent view of everything reachable from the root it loaded.
Persistent data structures make this the entire design, and \rcu's
build--publish--reclaim discipline (\cref{sec:buildpublish}) is the same idea in
the small --- which is exactly why an \rcu reader gets a coherent view of a
\emph{node} for free.

What copying does not do is scale to the update, and that is where this paper
began (\cref{sec:onestore}). When a change spans several words, copying a region
large enough to make it one pointer swap means paying for that region --- in
allocation, in footprint, and in reclamation --- on every update. The
alternative is to mutate in place.

Mutating in place is what ends the free lunch, and it is the regime every design
in \cref{tab:comparison} occupies, this one included. A reader can now arrive
while the structure is between states. Any property it must observe about a
mutation in flight has become a claim about what \emph{that reader} sees, and
only the reader can establish it. RLU establishes a snapshot by giving every
object a header that can name a newer copy, and having readers consult that
machinery when a writer's copies might be relevant to them; MV-RLU adds a
version chain for the reader to walk; STM has the reader log its reads and
validate them. Existence structures do not establish a snapshot at all --- their
guarantee is per-element --- but they pay in the same coin, giving every element
a field that can name a group and having readers load it. The property differs
and the mechanism differs; the location does not. It must be reachable
\emph{from the reader}, which is precisely why it lives in, or hangs off, the
elements themselves.

So there are two ways to spare the reader, and this design takes the second. It
declines to copy, because the cost of copying is what made the operation
expensive in the first place. And it declines to hand readers isolation, because
in a structure mutated in place isolation is a charge on the reader --- and the
reader is the one party \rcu exists to keep free.

\paragraph{Every design tests something; only this one reaches the test for
free.} The difference is structural rather than a matter of constant factors,
but it is narrower than "some designs check the read path and this one does
not". Every row of \cref{tab:comparison} checks it, this one included. What
differs is what the check costs to reach its operand.

Existence's fast path is one acquire load of a per-element field, then a branch
on it. RLU's fast path, for a thread with no writes of its own, never touches
the object header --- but it still loads a flag from its per-thread state, once
per dereference, and branches on that. MV-RLU reads the object header
unconditionally and branches, then may walk a version chain. None of these loads
is expensive in isolation; all of them are memory references, and in each case
the branch cannot resolve until its load does.

A \pseudotxn's marker occupies tag bits of the slot, so recognizing it is
\begin{center}
\code{((uintptr\_t) v \& tag) == tag}
\end{center}
where \code{v} is the value the reader's \code{rcu\_dereference} already
returned. That is a mask and a test on a register the reader was obliged to hold
anyway: no additional load and no additional cache line.

It is not free, and the residue is worth naming rather than rounding away. The
test is a conditional branch --- a couple of instructions, and one entry in the
branch predictor. It is hinted unlikely and is correctly predicted not-taken
whenever no commit is in flight, which is to say almost always; but a predicted
branch is not a free branch, and a reader in a tight traversal pays it on every
hop. So \cref{tab:comparison}'s \textbf{0} is zero \emph{loads}, not zero
instructions.

That is the honest shape of the claim, and it is still the shape we want: the
branch is common to every design in the table, so it is not where the comparison
is decided. What separates this design is that its test operates on a value the
reader already had, while every other design must first go and fetch the thing
it will test --- from a per-element field, a per-thread flag, or an object
header. The marker rides the dereference the reader had to perform regardless.

\paragraph{The unconditional cost is space.} Instruction counts are the smaller
half of the argument, because the state is the part that cannot be made
conditional. RLU's header is prepended to every object and existence's field
sits in every element, for as long as the structure lives, whether or not any
mutation is in flight. That is a permanent addition to the footprint of the data
the readers traverse --- and footprint is a read-side cost, paid in cache rather
than in instructions. A \pseudotxn adds no per-element state at all, so in the
steady state both the structure a reader traverses and the path it takes through
it are byte-for-byte what they would be with no engine present.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is where the rcu_head layout experiment belongs -- it isolates
% exactly this effect (segregated vs inline: roughly 2x read throughput at
% 1000 nodes) under our own control. It is not the same field, so present it
% as evidence for the MECHANISM (per-element growth costs read throughput),
% not as a measurement of existence or RLU. Overclaiming it as a direct
% measurement of either would be the same error as the RLU read-path claim
% we already had to retract. Cross-reference the evaluation paper.
% -----------------------------------------------------------------------

\paragraph{A pseudo-transaction charges the reader no memory.} The engine adds
no per-element state at all. Its marker lives in tag bits of the slot itself ---
the very word the reader had to load in order to proceed --- so recognizing it
costs no additional load and no additional line, only the predicted branch
accounted for above. And because the marker is transient, the structure a reader
traverses in the steady state is byte-for-byte the structure it would traverse
with no engine present.

\paragraph{The cost moves to the write side.} This is not a free lunch, and the
evaluation must not pretend otherwise. Where existence collapses an element's
entire membership into a single object it can flip once, a \pseudotxn names
individual slots, so its commit is as wide as the number of edges it changes. A
structure whose logical update touches many pointers --- a tower, a wide node
--- therefore commits many records where existence commits one. Commit width is
the scarce resource, and it is the axis on which this design is at its worst.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Say this plainly, and let the evaluation confirm it rather than apologize
% for it: the existence-vs-txn gap on a tower-structured workload should
% WIDEN with tower height, because width grows and existence's does not.
% That is a prediction the cost model makes, and it is testable with the
% engine's own commit-cost introspection. A validated cost model is a
% result; an unexplained loss is an embarrassment. They are the same
% numbers.
% -----------------------------------------------------------------------

\paragraph{The weakening is targeted, not residual.} We did not fail to provide
isolation; we declined to charge for it. The structures \rcu serves are
read-dominated by construction, so paying permanently, in every element, for a
snapshot those readers never requested spends the one resource they cannot
spare. A \pseudotxn removes exactly that charge and relocates the cost onto the
writers, who are by hypothesis the minority.

One qualification belongs here rather than in a footnote, because the argument
above is otherwise too flattering. What RLU buys with its charge is strictly
more than what we provide: a coherent snapshot, where we give only monotone
resolution (\cref{prop:monotone}). "We declined to charge for it" is therefore
an honest description of the trade only if it travels with what the charge
bought. A reader who needs a snapshot cannot take this engine and economize ---
they need a different engine. The comparison holds where the guarantee classes
match, as they do against existence structures (\cref{sec:costcomparison}); a
comparison against RLU crosses them (\cref{sec:related}).

% ---------------------------------------------------------------------------
% GUARDRAIL (do not delete in an editing pass).
%
% The paragraph above is the honest asterisk on Sec. 3.7's whole argument, and
% it must stay IN THE TEXT rather than become a footnote. RLU's tax buys
% strictly MORE than ours does -- a coherent snapshot, where we give "never
% new-then-old" -- so a comparison against RLU crosses guarantee classes and
% must say so. The comparison against existence structures is class-matched and
% needs no such qualification. Sec. 9 states the same asterisk for the
% throughput comparison specifically; both are needed, because Sec. 3.7 is
% where a reader forms the impression that the trade was free.
% ---------------------------------------------------------------------------

% ===========================================================================
\section{The Single-Writer Flip-Latch}
\label{sec:fliplatch}

\subsection{The intermediate-state problem}
\label{sec:intermediate}

A structural \rcu mutation often has to re-point many slots at once. Re-parenting
a set of live nodes must rewrite each node's back-pointer; a reader walking
several of them must not see a mix of old and new targets. Re-pointing the slots
one at a time cannot offer that: each store is individually atomic, but between
them the set is part-old and part-new, and a reader traversing the set observes
exactly that.

The conventional remedy is to drain the readers --- a grace period around the
update, so that no reader spans the inconsistent window. It is correct, and it
is expensive: the writer waits out every reader in the system to publish a
change none of them were looking at.

\subsection{Mechanism: one level of indirection}
\label{sec:swmechanism}

The flip-latch removes the intermediate states rather than hiding them behind a
drain, and it does so with one level of indirection
(\cref{fig:fliplatch}).

Each slot in the set is made to hold a \emph{tagged} pointer to a small
\emph{proxy} rather than the target directly. A proxy carries
$\{\mathit{old},\mathit{new}\}$ and a pointer to a shared \emph{flip group}
carrying one boolean \emph{selector} word. Resolving a proxy returns
$\mathit{old}$ while the selector is $0$, and $\mathit{new}$ once it is $1$.

Everything follows from one property: \textbf{every proxy in a group reads the
same selector.} A single store flipping that selector therefore switches the
whole set at once, and at every instant every proxy resolves consistently ---
all old, or all new. There is no instant at which the set is part-old and
part-new, because there is only one word that decides, and one store that
changes it.

\input{fig-fliplatch}

This is the commit step of \cref{def:pseudotxn}, realized. The record set is the
set of proxies to be installed; it is frozen when installation begins, because
from that point the writer is placing proxies for exactly those records and no
other may join. Installation itself publishes nothing --- the selector is still
$0$, so every proxy resolves to old --- and the commit step is the lone selector
store that follows. That single store, $0 \to 1$ and never written back, is also
why \cref{prop:monotone} holds.

\subsection{Why exclusion is required}
\label{sec:exclusion}

The flip-latch is legal \emph{only} under writer mutual exclusion. This is a
precondition of the mechanism, not a caveat on it.

The engine has no install-time CAS, no conflict detection, and no abort. Those
are precisely the parts that would let two writers discover each other, and they
are absent by design --- their absence is what makes this engine cheaper than the
multi-writer one. The consequence is unforgiving: two writers racing on one slot
simply corrupt it, and a default build says nothing.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Do not soften this sentence. It is the whole contract, and an embedder who
% misreads it writes a corruption that no test in a default build will
% surface.
% -----------------------------------------------------------------------

Because the failure is silent, the engine ships a runtime validator
(\code{-DURCU\_TXN\_SW\_EXCL\_VALIDATE}) that aborts the process on a violation.
It is off by default and costs nothing when off --- the handle does not even
carry the owner field. It claims the two things that exist to be claimed: the
\emph{handle}, whose owner is the thread that initialized it (catching a
\pseudotxn handed between threads mid-flight), and the \emph{slot}, which is
what two racing writers actually share and so is where the violation is really
visible. In a correct single-updater program a slot being recorded or parked
cannot already hold a parked proxy.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is worth stating as a design lesson rather than an implementation
% note: the contract is unenforceable at zero cost, so the engine makes the
% check opt-in and free-when-off rather than pretending the contract is
% safe. Ties to the series-wide honesty point: memory safety is
% unconditional, atomicity is conditional on embedder contracts. Cross-
% reference Sec. 7.
% -----------------------------------------------------------------------

\subsection{Lifecycle: build, install, commit, settle}
\label{sec:lifecycle}

\begin{enumerate}
\item \textbf{Build} the new structure where readers cannot reach it.
\item \textbf{Install.} The record set is frozen at this point: these are the
      slots that will change, and no further record may join. Initialize the
      group, then for each slot initialize a proxy with its
      $\{\mathit{old},\mathit{new}\}$ targets and point the slot at the tagged
      proxy. The selector is $0$, so this is \emph{transparent to readers}: they
      still resolve to old. Installation is not a partial commit; it is not
      observable at all.
\item \textbf{Commit.} One release store, $0 \to 1$. Every proxy now resolves to
      new, atomically. This is the linearization point
      (\cref{prop:lp}).
\item \textbf{Settle.} Rewrite each slot from the tagged proxy back to the
      direct new target. This is idempotent for readers, since the proxy already
      resolves to new --- so settle changes no observable value and is not a
      linearization point either. Then \code{call\_rcu()} the proxies and the
      group.
\end{enumerate}

The shape is exactly build~$\rightarrow$~publish~$\rightarrow$~reclaim
(\cref{sec:background}), with the commit as the single publish. Steps 2 and 4
bracket it symmetrically: install makes the indirection reader-visible without
changing any value, and settle removes it without changing any value.

The claim that only one of these four steps is observable is easy to assert and
worth checking. \Cref{tab:lifecycle} tracks a single transacted slot $s$ through
the lifecycle, alongside what a concurrent reader resolving $s$ actually
obtains.

\begin{table}[t]
\centering
\caption{One transacted slot $s$ through the lifecycle. The reader's column
changes exactly once. Install and settle both rewrite $s$ --- they are stores to
a word readers are concurrently loading --- yet neither changes what resolving
$s$ yields, which is why the commit is the only linearization point.}
\label{tab:lifecycle}
\begin{tabular}{@{}llll@{}}
\toprule
Step & $s$ holds & Reader resolving $s$ obtains & Linearization point? \\
\midrule
(before)  & $o$, directly            & $o$ & --- \\
Build     & $o$, directly            & $o$ & no: new nodes unreachable \\
Install   & tagged proxy, \code{sel}$=0$ & $o$ & no: resolved value unchanged \\
Commit    & tagged proxy, \code{sel}$=1$ & $n$ & \textbf{yes} \\
Settle    & $n$, directly            & $n$ & no: resolved value unchanged \\
(after)   & $n$, directly            & $n$ & --- \\
\bottomrule
\end{tabular}
\end{table}

The third column is the mechanism. It reads $o, o, o, n, n, n$: one transition,
at the commit, no matter how many slots the set contains or how long the writer
takes over the other three steps. A reader cannot tell installed-but-uncommitted
from untouched, and cannot tell settled from committed-but-unsettled. Only the
selector store is visible, and it is one store.

\paragraph{The wrinkle, which is the contract in disguise.} Across the install
window a proxy resolves to the old target \emph{recorded at install time} rather
than to whatever the slot happens to hold. Under exclusion those are the same
thing --- no one else writes $s$, so the recorded $o$ stays the true $o$ --- and
the row for Install above is exact.

Remove exclusion and that row becomes a lie in the most inconvenient way
available. A second writer's store to $s$ during the window is not detected and
not lost-but-noticed: the proxy goes on publishing the $o$ it recorded, so
readers resolve to a value that is no longer what the slot was set to, and the
subsequent settle overwrites the intruder's store with $n$. Nothing traps, and
readers are handed a stale value \emph{as though it were current}. This is the
same fact as \cref{sec:exclusion} viewed from the reader's side, and it is worth
seeing from both: the \fliplatch's cheapness comes from having no install-time
check, and an engine with one --- the \mw engine's install
compare-and-swap --- would have refused the install instead.

\subsection{Progress}
\label{sec:swprogress}

Under exclusion there is nothing to contend for, and the progress story is
correspondingly short: a commit cannot fail. The engine has no abort because it
has no contention to abort on.

This is worth stating plainly because the shared status enumeration carries an
\ABORT value, and its presence invites the opposite conclusion. That value
exists so that an embedder can be written once against both front-ends: only the
concurrent engine ever returns it, and the single-updater, having no contention,
never does.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is the migration-shape argument and it pays off in the companion
% paper: the *_prepare forms and the shared status enum are what let an
% embedder move from this engine to the multi-writer one mechanically.
% Forward-reference it, but do not develop it here -- P1 should not depend
% on P2 to be complete.
% -----------------------------------------------------------------------

\subsection{What the embedder owes}
\label{sec:embedderowes}

The engine is deliberately incomplete: it supplies the latch, and leaves three
obligations with the structure that embeds it.

\begin{itemize}
\item \textbf{Tagging.} The embedder chooses how a proxy pointer is marked so
      its readers recognize one --- a spare low or high bit, or a reserved type
      code in an already-tagged pointer --- and routes resolution through the
      engine's accessor. \Cref{sec:tags} takes this up: it is where the engine's
      zero-added-state claim is actually cashed.
\item \textbf{Allocation.} Proxies and the group are the embedder's to allocate,
      with whatever alignment its tagging scheme demands --- often as a single
      backing block with one \code{rcu\_head}.
\item \textbf{Reclamation.} After settle has unpublished every proxy, the
      embedder \code{call\_rcu()}s them.
\end{itemize}

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Frame this as a design position, not a shortfall: the engine cannot know
% the embedder's pointer encoding, so a general engine that OWNED the
% tagging would have to add a field -- which is the very cost this design
% exists to avoid. The obligations are the price of adding no per-element
% state, and that trade is the paper's thesis. Do not let this read as "and
% the user has to do the hard part".
% -----------------------------------------------------------------------

% ===========================================================================
\section{Slots and Tags}
\label{sec:tags}

\Cref{sec:whyweakening} claimed that this engine adds no per-element state and
costs the reader no additional load. This section is where that claim is
cashed, and where its price is stated: the engine has nowhere of its own to put
the marker, so it must borrow bits from the embedder's pointer encoding.

\subsection{The tag contract}
\label{sec:tagcontract}

A parked slot must hold something a reader can recognize as a proxy and
distinguish from every legitimate value the slot could otherwise hold. The
engine takes those bits from the slot itself:

\begin{itemize}
\item a parked slot's value is \code{(record\_address | tag)};
\item the record is recovered as \code{(value \& \textasciitilde tag)};
\item a value is this record's proxy iff \code{(value \& tag) == tag}.
\end{itemize}

Which imposes a contract on the embedder, and it is unforgiving:
\textbf{every value the slot may legitimately hold must satisfy
\code{(value \& tag) != tag}.} A single value that happens to carry the tag
pattern is indistinguishable from a proxy, and a reader will follow it into a
record that does not exist.

Two further obligations follow. Every write to a transacted slot must go through
the engine: one raw store breaks atomicity for every participant, not only the
writer that made it. And every read must resolve through the engine's accessor,
since a bare load of a parked slot yields a tagged record address rather than a
value.

\subsection{Slots need not hold pointers}
\label{sec:nonpointer}

Nothing in \cref{sec:tagcontract} mentions pointers. The contract constrains
\emph{values}, and the engine transacts \emph{words} --- what a word means is
the embedder's business. A transacted slot may hold any word-sized value that
keeps the tag bits clear.

For integers the discipline is to store them shifted left by one, so bit~0 stays
clear and a word carries one bit fewer than its width --- 63 data bits on a
64-bit word. A bitmap is the limiting case: it is \emph{all} data, with no
spare bits of its own to donate, so it spends bit~0 as the tag and lays logical
bit $i$ at physical bit $i+1$. Nothing about the engine changes, and no sentinel
value has to be carved out of the data range. A zero-filled region is a valid
empty bitmap --- all bits clear, bit~0 clear --- so \code{calloc}'d or
demand-zero storage needs no initialization.

This matters beyond generality, and it is the sharpest form of the argument in
\cref{sec:whyweakening}. A design that attaches metadata to elements needs an
\emph{element} to attach it to. A bare word of bitmap is not an object: it has
no header to prepend and no field to add, so making it transactable under such a
scheme means wrapping it in one --- paying an object's worth of metadata to
transact a word's worth of data. This engine spends one bit.

The transacted bitmap is also the plainest case for the composition of
\cref{sec:swcomposition}. Because 63~logical bits share one physical word,
flipping two different bits records two edits on the \emph{same} slot; they
compose into one commit only because the recording store fuses same-slot edits
rather than appending them. The bitmap is where read-your-own-writes stops being
a list-specific convenience and becomes the general rule.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% State this carefully. The fair version is about GRANULARITY, not a claim
% that RLU/existence "cannot" do it: you can always wrap a word in a struct
% and give the struct a header. The point is that the wrapping IS the cost,
% and it is ~100% overhead for a bare data word, whereas the tag is one bit.
% Do not overstate into "impossible" -- it is trivially refutable and the
% honest version is already strong.
% -----------------------------------------------------------------------

\subsection{One commit, two tag widths}
\label{sec:mixedtags}

The bitmap of \cref{sec:nonpointer} keeps its tag as narrow as possible ---
bit~0 alone --- for a reason that is not frugality. It is so that a bitmap word
can share a commit with pointer slots that need a \emph{wider} tag.

That is \cref{sec:pertag}'s argument in the concrete: an occupancy bit flipping
atomically with the child-pointer store it accompanies, where the bitmap word is
tagged with bit~0 and the pointer slot is tagged with a reserved nibble because
its own encoding has already spent the low bits. One \pseudotxn, two tag widths,
resolved per record. A per-translation-unit tag could not express it.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is the worked example Sec. 5.2 needed and it costs nothing: it is
% statable entirely in terms of a bitmap and a typed pointer, with no
% dependency on the deferred consumer. Use it INSTEAD of the fractal trie,
% not alongside.
% -----------------------------------------------------------------------

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% RESOLVED (pinned cefb0414): the single-writer bitmap wrapper now exists,
% include/urcu/rcu-txn-sw-bitmap.h. Both questions the old PENDING note flagged
% are answered in source: (1) it keeps the narrow bit-0 tag (so Sec. 5.3's
% mixed-width example holds -- a bit-0 bitmap word composing with a
% nibble-tagged pointer in one commit); (2) rank()/select() read across several
% words and are per-word HINTS, not snapshots -- they hold no proxy and resolve
% each word against its own moment, so a multi-word scan can straddle a commit.
% That is the general "a reader is not a transaction" property (Sec. 3.4), not a
% bitmap-specific caveat, and needs no separate treatment. The MCAS twin offers
% a guarded multi-word snapshot; the sw engine deliberately does not.
% -----------------------------------------------------------------------
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Be explicit that this is the cost side of the Sec. 3.7 bargain. We pay no
% space and no load; we pay in a CONTRACT the embedder must honour and the
% engine cannot check for free. That is the same shape as the exclusion
% precondition (Sec. 4.3) and the same shape as the series' honesty point:
% memory safety unconditional, atomicity conditional. Sec. 7 collects these.
% -----------------------------------------------------------------------

\subsection{Why the tag cannot be fixed}
\label{sec:pertag}

The obvious design is a single tag, chosen once by the engine --- bit~0, say,
free in any aligned pointer. It is also the design that
\cref{prop:crossstructure} rules out.

Bit~0 is free only in structures that have not already spent it. A structure
whose pointers encode a type or a mark in their low bits has not; for such a
structure, a bit-0 proxy \emph{aliases a real value}, and the contract of
\cref{sec:tagcontract} is violated by construction. Such a structure can still
be transacted --- it simply has to nominate a different pattern, one that no
value of its own can carry, perhaps a reserved low nibble rather than a single
bit.

That alone would argue for a per-embedder tag, which a compile-time constant
could supply. What forces the tag \emph{per record} is cross-structure
atomicity. One \pseudotxn may span structures with different encodings: the
whole point of \cref{prop:crossstructure} is that a single commit can publish a
node into one structure and splice it into another. Those slots have different
free bits, so \emph{one transaction needs two tags at once}. A
per-translation-unit macro cannot express that; the tag has to travel with the
record. \Cref{sec:mixedtags} gives the concrete case.

So each record carries its own \code{proxy\_tag}, supplied when the record is
added, and heterogeneous slots share one engine, each resolved through its own
tag.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is the tightest argument in the paper: Property 4 (cross-structure
% atomicity) FORCES the per-record tag. It is not a convenience or a
% generalisation-for-its-own-sake -- a per-TU tag would make Property 4
% unimplementable for heterogeneous embedders. Lead with that, not with
% "flexibility". Do NOT use the fractal trie as the worked example here --
% it is deferred to a later paper and P1 must not depend on it. The hazard
% is fully statable in the abstract: "a structure whose pointers already
% encode a type in their low bits". The engine header names the trie's 0xF
% nibble as its motivating case; we can describe the shape without the
% consumer.
% -----------------------------------------------------------------------

\subsection{Storing the tag, not a tagged pointer}
\label{sec:tagnotptr}

A record stores the \emph{tag}, not a pre-tagged pointer, and the proxy is
formed from the record's current address at install time. That distinction looks
pedantic and is not: it is what makes a descriptor safe to grow.

A transaction that outgrows its descriptor must move its records. Had each
record cached a proxy value computed from its old address, every such value
would be stranded by the move --- pointing at storage the records no longer
occupy. Because the record stores only the tag, the tag travels with the
record's content through the move, and the installer re-derives the proxy from
the record's final address. Nothing is stranded, and growth needs no fixup pass.

\subsection{The alignment budget}
\label{sec:alignment}

The tag's width is bounded by how many low bits of a record address are
guaranteed clear. Records are 16-byte aligned, so an embedder has four bits to
choose from: enough for a single bit, or for a reserved nibble.

The alignment is established deliberately rather than inherited. The descriptor
is allocated with \code{posix\_memalign()} to a 16-byte boundary, and the record
array's offset and stride are both multiples of 16, so every record address ---
including every inline record --- has its low four bits clear. This holds
\emph{portably}: relying on \code{malloc} would rely on an allocator's habits
rather than a guarantee, and \code{max\_align\_t} is only 8 bytes on some 32-bit
ABIs.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Small but worth one sentence: "we asserted the alignment we need rather
% than inheriting whatever the allocator happened to give us" is exactly the
% kind of detail an enabling disclosure needs, and exactly the kind a normal
% paper drops. Keep it.
% -----------------------------------------------------------------------

\subsection{What the tag buys}
\label{sec:tagbuys}

The tag is the whole of the engine's presence in a reader's world. There is no
header, no side table, no per-element field --- the marker occupies bits of a
word the structure already had and the reader already loads. That is why
\cref{tab:comparison}'s last row reads zero in the loads column, and it is why
the structure a
reader traverses in the steady state is byte-for-byte the one it would traverse
with no engine linked at all.

The price is that the engine cannot own the encoding. It does not know which
bits a given structure has left, so it cannot choose for the embedder ---
\cref{sec:embedderowes}'s obligations are the direct consequence. An engine that
owned the tagging would have to add a field to say what it owned, which is
precisely the cost this design exists to avoid.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Close the loop with Sec. 3.7 explicitly: the obligations are not a
% shortfall, they are the mechanism by which the cost is avoided. An engine
% that took the decision away from the embedder would have to pay for the
% privilege in exactly the coin (per-element state) the design refuses to
% spend.
% -----------------------------------------------------------------------

% ===========================================================================
\section{Transacted Data Structures}
\label{sec:wrappers}

\subsection{The gap: a list with no reverse walk}
\label{sec:rculistgap}

Userspace \rcu's doubly-linked list, \code{rculist.h}, ships forward iterators
and no RCU-safe reverse walk. That is not an oversight, and the reason is worth
following, because it is the concrete form of the problem this engine solves.

A doubly-linked list has two chains, but \code{rculist} publishes only one. In
\code{cds\_list\_del\_rcu()} the forward edge is committed by a single atomic
store to \code{prev->next}. The back edge is a plain assignment: no
\code{rcu\_assign\_pointer}, no release ordering --- and correspondingly, the
\code{prev} pointers are never \code{rcu\_dereference}d. A backward walk
therefore has no acquire to pair with and no atomic handoff, and can read a
\code{prev} value inconsistent with the live forward chain.

The semantic consequence is worse than the ordering one. A removed node is a
ghost, reclaimed only after a grace period. A reverse walker can keep stepping
onto a node the forward chain has already dropped, for as long as that ghost
survives --- so \textbf{the \code{prev} view trails the \code{next} view by up to
a grace period}. There is no barrier to sprinkle that fixes this: the two chains
are updated by two stores, and nothing makes them one event.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is the paper's best motivating example and it should be in Sec. 1
% too, in one sentence. It is a real, shipped, widely-deployed library that
% OMITS AN OPERATION because it cannot publish two edges atomically. Nobody
% has to be persuaded the problem is real -- the absence of the API is the
% evidence.
% -----------------------------------------------------------------------

\subsection{Two edges, one flip}
\label{sec:swlist}

Every structural change to a doubly-linked list re-points exactly two
reader-visible edges, one forward and one backward:
\begin{center}
\begin{tabular}{@{}ll@{}}
insert $E$ between $P$ and $N$: & \code{P->next}: \edit{N}{E} \quad and \quad \code{N->prev}: \edit{P}{E} \\
delete $E$ between $P$ and $N$: & \code{P->next}: \edit{E}{N} \quad and \quad \code{N->prev}: \edit{E}{P} \\
\end{tabular}
\end{center}

Two edges is exactly the case \cref{sec:swmechanism} handles. Both slots
transiently hold tagged proxies sharing one selector word; one release store
flips the selector and switches both edges together. At every instant the
forward and backward chains describe the same list, and the reverse walk that
\code{rculist} cannot offer becomes ordinary.

\subsection{The coherence guarantee, and its scope}
\label{sec:coherence}

\begin{property}[Mutual inverses]
\label{prop:coherence}
For any two nodes that are both members of the list at the instant they are
resolved, the links are mutual inverses: \code{a->next == b} implies
\code{b->prev == a}. No operation exposes a window in which one direction sees
the change and the other does not.
\end{property}

This does not freeze the list for a long-lived reader --- nothing here does, and
\cref{sec:nonguarantees} said so. A reader may still observe an operation that
commits during its walk. What is guaranteed is that whatever it observes is
internally coherent in both directions \emph{at each step}.

``At each step'' is the precise scope, and it is \cref{sec:monotone}'s
dependency-chaining caveat in concrete form. Coherence is a property of a single
resolution; carrying it across two resolutions needs an ordering edge between
them. A dependency-chained walk --- next-then-next, or a \code{prev} hop taken
\emph{from} the node the previous load returned --- has one on every supported
architecture, as does any build using the default C11 dereference. A reader
comparing two independently-reached slots, a cached pointer or a \code{prev} and
a \code{next} read from nodes it did not chain between, must supply its own
acquire.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Do not bury this. It is the same precondition as Property 5's scope
% paragraph, and stating it twice -- abstractly in Sec. 3.5, concretely here
% -- is correct rather than repetitive, because this is where an embedder
% will actually trip over it.
% -----------------------------------------------------------------------

\subsection{The hash list: one visible edge, two recorded}
\label{sec:swhlist}

The hash-bucket sibling is shaped like the kernel's hlist: the head is a single
8-byte pointer, so a bucket table costs half what sentinel-node heads cost.

Its backward link is the elegant part. A node's \code{pprev} is a pointer to the
\emph{slot that names it} --- \code{\&prev->next}, or \code{\&head->first} ---
rather than a pointer to the previous node. So the head's single pointer is just
another \code{next} slot, and \textbf{no operation carries a head special
case}: first-position insert and delete are the general code path with a
different slot address.

Readers walk forward only, so each structural operation has exactly one
reader-visible edge; \code{pprev} is writer-only bookkeeping. Both edges are
nonetheless recorded into the flip transaction, for two reasons that have
nothing to do with reader visibility. A mutation becomes all-or-nothing under an
allocation failure. And it composes: through the \code{*\_prepare} forms, a
single-edge hlist edit can be folded into a larger cross-structure flip ---
which is \cref{prop:crossstructure} being cashed, and the reason even an
operation that needs no atomicity for its own sake is still expressed as a
\pseudotxn.

\subsection{Composing edits: read-your-own-writes}
\label{sec:swcomposition}

Folding several edits into one commit raises a hazard that has nothing to do
with concurrency, and everything to do with a writer reading state its own
earlier edit has already changed. Consider deleting two adjacent nodes,
$P \rightarrow E_1 \rightarrow E_2 \rightarrow N$, in one \pseudotxn. Removing
$E_1$ records \code{P->next}: \edit{E_1}{E_2} and \code{E2->prev}:
\edit{E_1}{P}. Removing $E_2$ must now record \code{P->next} again --- but only
if it \emph{sees} that \code{P->next} is already pending. If it re-reads the
live structure instead, it believes $E_2$'s predecessor is still $E_1$, records
\code{E1->next}: \edit{E_2}{N} on a node that is about to vanish, and the commit
publishes a list pointing through deleted nodes. Nothing here is a race; one
writer, editing in program order, has torn its own commit.

The engine closes this the way its concurrent sibling does, with
\emph{read-your-own-writes}. Two accessors form the pair: a load that returns a
slot's pending new value if this \pseudotxn has already recorded one and its
committed value otherwise, and a recording store that \emph{fuses} a second edit
of a slot onto the existing record --- keeping the committed old, advancing the
new --- rather than appending a duplicate. A transaction holds at most one record
per slot. The wrappers read every neighbour link through the load and record
through the fusing store, so the second delete above reads the \emph{pending}
\code{P->next}, chains onto that record, and the commit unlinks both victims and
leaves $P \leftrightarrow N$ coherent. This is the same mechanism, and the same
guarantee, as the concurrent engine's, reached through the same calls --- which
is what lets an embedder written against one move to the other without rewriting
its commit handling.

Two boundaries remain, and both are narrower than the old ``slot-disjoint
only'' rule they replace.

\emph{The caller must name the nodes it means.} Read-your-own-writes reconciles
the \emph{links} an edit records, but it cannot re-run the caller's search: a
traversal made with the ordinary \rcu accessors sees the committed structure,
not this \pseudotxn's pending one, so a caller that walks and edits in the same
bracket can still hand an edit a node its own earlier edit has displaced. The
discipline is to identify the victims first and then edit, not to interleave the
two.

\emph{Declaring a write set disjoint is a promise the engine keeps you to.}
Read-your-own-writes costs a find per recorded edit, which an embedder whose
slots are distinct \emph{by construction} can skip by declaring the handle
disjoint --- the single-writer analogue of the concurrent engine's identical
option. The declaration is a contract on slot \emph{addresses}: a bitmap, whose
63~logical bits share one word, satisfies it only for edits the caller knows
fall in distinct words. Break the promise and the fused reconciliation is
skipped, the earlier same-slot edit is lost, and the commit reports success ---
which is why a debug build
(\code{-DURCU\_TXN\_SW\_DEBUG\_DISJOINT}) traps a violated declaration, and why
the declaration should be made only where disjointness is structural rather than
merely observed.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This section was rewritten when the sw engine gained read-your-own-writes
% (urcu-txn-dev 10f22a76, d47aaae8, cced8fb0; pinned at cefb0414). BEFORE that,
% the sw engine had no transactional loads: composition was slot-disjoint ONLY,
% and adjacent deletes silently corrupted -- which was true at the previous pin
% (ec621268) and is the behaviour earlier drafts described.
%
% Do NOT reinstate the old "the concurrent engine's RYW is what it is FOR / the
% reason it has transactional loads at all" framing anywhere (it was in this
% NOTE and in the Sec. 10 P2 hook). It is now FALSE: BOTH engines have RYW, so
% RYW is part of the SHARED programming model (P3), not P2's differentiator.
% What still distinguishes the concurrent engine is the machinery exclusion
% removes -- the install CAS, conflict detection, abort, helping -- for MULTIPLE
% writers. RYW is orthogonal to that and now lives on both sides.
% -----------------------------------------------------------------------

\subsection{Composing with reclamation}
\label{sec:reclaim}

The wrappers inherit the discipline of \cref{sec:background} without amendment,
because the commit is the publish step.

A node's payload is initialized before the commit that links it, and the commit
is the release edge, so the payload is published safely. Nodes that a committed
\pseudotxn unlinked were reader-visible, so they are freed by
\code{call\_rcu()} after a grace period. Nothing an aborted attempt allocated
was ever published, so it is freed immediately --- but only what \emph{that
attempt} allocated, never anything already reachable.

The proxies themselves are subject to the same rule, and it is why settle exists
(\cref{sec:lifecycle}): a proxy is reader-reachable while installed, so it may
be reclaimed only once settle has rewritten every slot that named it, and then
only after a grace period.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% RESOLVED -- do not add the "no re-link before a grace period" trap here.
% It is MULTI-WRITER ONLY and belongs in the companion paper. Checked: rcu-
% txn-list.h (the concurrent list) warns that a deleted node must not be re-
% linked before a grace period, because "the deleting transaction may still
% hold a parked proxy in that node's next slot, awaiting its settle. A plain
% store into a proxied slot is then overwritten when that settle converts
% the proxy, and the write is silently lost -- a permanently incoherent
% edge." That hazard needs a PEER transaction holding an unsettled proxy.
% Under exclusion the sole writer's transactions are sequential and settle
% completes (lifecycle step 4) before the next begins, so no proxy of a
% previous transaction survives into the next one. Confirmingly, rcu-txn-sw-
% list.h carries no such warning: it mentions grace periods only for
% reclaim. Stating it here would import a multi-writer hazard into a single-
% writer paper and imply a restriction the sw engine does not have.
% -----------------------------------------------------------------------

% ===========================================================================
\section{Limitations}
\label{sec:limitations}

The engine's limits divide cleanly in two, and the division is the useful part.
Some are \emph{properties of the model} --- they cannot be violated, only
disliked. The rest are \emph{contracts on the embedder} --- they can be
violated, and violating them is expensive.

\subsection{What holds unconditionally}
\label{sec:unconditional}

One thing is worth stating positively first, because it is the guarantee that is
usually hardest to get in this design space and it costs nothing here.

The engine is indifferent to \emph{slot-value} A-B-A. Resolution goes through
the record's own status rather than through the slot's plain value, so the
identity that matters is the descriptor's, not the value's. Any slot value may
recur --- a next-pointer cycling $B \rightarrow X \rightarrow B$ with $B$ a live
successor that is never freed, a counter revisiting a number, an embedder's own
reuse --- and the recurrence is inert. \rcu prevents \emph{address} reuse, but
not \emph{value} recurrence; this engine does not need it to.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Be precise about the scope of "unconditional" here, because it is easy to
% overstate and the overstatement is load-bearing. The engine's own words
% are "memory safety, which holds unconditionally here" -- and the sentence
% is about SLOT VALUES. It is NOT a claim that memory safety survives a
% violated contract. It cannot be: Sec. 7.2's tag contract, violated, makes
% a reader follow a non-proxy value as a record address. Do not write
% "memory safety is unconditional, atomicity is conditional" as a blanket --
% it is wrong in the direction that flatters us, which is the worst
% direction.
% -----------------------------------------------------------------------

\subsection{The contracts, and what violating them costs}
\label{sec:contracts}

The engine requires four things of its embedder. None is checked in a default
build.

\begin{description}
\item[Writer exclusion.] The mechanism has no install CAS, no conflict
  detection and no abort (\cref{sec:exclusion}). Two writers racing one slot
  corrupt it.
\item[Free tag bits.] Every value a transacted slot may legitimately hold must
  satisfy \code{(value \& tag) != tag} (\cref{sec:tagcontract}).
\item[Correct composition, and no raw access.] Edits that touch the same slot
  must be composed through the read-your-own-writes path, which fuses them
  (\cref{sec:swcomposition}); the blind fast path and any handle declared
  disjoint instead require pairwise-distinct slots. Every write to a transacted
  slot must go through the engine and every read through its accessor.
\item[Deferred descriptor reclaim.] A descriptor reachable through a slot is
  reclaimed only after a grace period, because a reader may dereference it.
\end{description}

What a violation costs is worth stating exactly, because it is more than
atomicity. A value that accidentally carries the tag pattern is
indistinguishable from a proxy, and a reader will untag it and dereference it as
a record: that is a wild pointer, not a torn read. Racing writers corrupt slots
whose values readers then follow. \textbf{These contracts underwrite memory
safety, not merely atomicity} --- and that is the honest shape of the safety
argument.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is the series' central honesty point and Sec. 1 should not oversell
% around it. The July review of this engine concluded that every major
% finding sat at the CONTRACT layer rather than the mechanism layer -- doc
% text with API force steering embedders into corruption. That is a real
% result about this class of design and worth saying: the mechanism is the
% easy part.
% -----------------------------------------------------------------------

\subsection{Contracts the engine cannot check for free}
\label{sec:checkability}

The contracts of \cref{sec:contracts} are not equally checkable, and the engine
is candid about which is which rather than uniformly optimistic.

Exclusion has an opt-in runtime validator
(\code{-DURCU\_TXN\_SW\_EXCL\_VALIDATE}) that aborts on violation and costs
nothing when off --- the handle does not even carry the owner field. It claims
the handle (catching a \pseudotxn handed between threads) and the slot (where
two racing writers actually collide).

The tag contract has no such check, and cannot cheaply have one: verifying it
would mean inspecting every value ever stored in a transacted slot, which is the
per-write cost the design exists to avoid. It is discharged by the embedder's
encoding, by construction or not at all.

A disjoint declaration (\cref{sec:swcomposition}), if false, is the quietest
violation of all: the engine skips the reconciliation the declaration promised
was unnecessary, drops the earlier of two edits that land on one slot, and
commits success. It is the one contract with a dedicated opt-in trap
(\code{-DURCU\_TXN\_SW\_DEBUG\_DISJOINT}), though even that catches only the
recording side; a default build discharges the declaration by construction or
corrupts quietly.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Frame the opt-in validator as the design position it is: the contract is
% unenforceable at zero cost, so the engine offers a check that is free when
% off rather than pretending the contract is safe or taxing everyone to
% police it. That is the same trade as the rest of the paper -- pay nothing
% by default, and be explicit about what that costs.
% -----------------------------------------------------------------------

\subsection{Properties of the model, not defects}
\label{sec:modellimits}

The remaining limits cannot be violated because they are not promises. They are
what the model \emph{is}.

\textbf{No reader snapshot.} A reader is not a transaction of any kind
(\cref{sec:nonguarantees}); a traversal spanning slots may straddle a commit. An
embedder that needs a coherent multi-slot read builds it on top, and ongoing
work in this series develops more than one way. A \emph{seqcount} around the
commit lets a reader notice that a mutation intervened and retry. A plain
seqcount reader is only obstruction-free --- a sustained stream of writers can
starve its retry indefinitely --- but the grace period repairs exactly that. If
the counter's next in-place advance is withheld until the grace period following
the previous one has elapsed, no reader's sampling window can span two advances,
so a reader retries \emph{at most once}: a bound that is constant in the number
of writers, which is a wait-free read. One bit of counter suffices for it, and
the cost moves to the writer, which admits one in-place update per grace period
per node and copies the rest. A \emph{validating transaction} is the other
direction: it folds the reader's loads into the commit's conflict set --- the
loads ``act as writes'', accepted only if every slot still holds what it
returned --- torn-free, but it aborts and retries under contention. These are
points on one curve trading write-side work for read-side progress, with
copy-on-write at the far end; the base model fixes none of them, and that it
declines to is the deliberate weakening of \cref{sec:whyweakening}, not a
shortfall.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% The snapshot approaches are ONGOING work, deliberately NOT pinned. Provenance:
%   - seqcount reader + the single-bit GP-gated latch: design/rcu-txn-blob.md in
%     efficios-trie-benchmark; the latch is urcu-txn-dev 2a978059 (UNPUSHED as of
%     the pin cefb0414). The validating-transaction reader is the concurrent
%     engine's urcu_txn_load_validate() (rcu-txn.h).
% Hint only -- do not describe either as shipped, do not pin.
%
% CORRECTION (do not regress): an earlier draft, following the OLDER commit
% 98c7f2f7 ("seqcount reads are not wait-free"), claimed BOTH approaches forfeit
% wait-free reads and cast single-pointer COW as the only wait-free way to
% snapshot many words. Mathieu flagged this as wrong and it is: the single-bit
% GP-gated latch gets BOTH torn-free AND wait-free. Mechanism (blob design note,
% "Single-bit latch: wait-free reads for real"): gate the latch so the next
% in-place flip waits for the grace period after the previous flip, so no
% reader's window spans two flips; the reader then retries AT MOST ONCE -- a
% constant independent of writer count, hence genuinely wait-free, not merely
% bounded-by-a-large-constant. Cost moves entirely to the write side: <= 1
% in-place update per GP per node, recompact (COW) the rest. So there is NO
% wait-free-XOR-torn-free trilemma; COW is one point on a curve, not the only
% wait-free corner. Do not reinstate the trilemma framing.
% -----------------------------------------------------------------------

\textbf{Value-based, not stability-based.} A record's old is the value the slot
must hold \emph{at} the commit point, not one it must have held throughout. An
operation whose correctness requires \emph{detecting} an away-and-back change
must carry its own version in the word. Under exclusion this has no teeth. The
one thing that could invalidate a recorded old --- an earlier edit in the
writer's own bracket --- is exactly what read-your-own-writes reconciles
(\cref{sec:swcomposition}), and nothing outside the bracket can change the slot
at all. The distinction earns its keep in the concurrent engine, where a peer
writer can move a slot away and back between a read and the commit; here it is
inert.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% UPDATED when the sw engine gained read-your-own-writes (pinned cefb0414). The
% earlier draft unified "value-based not stability-based" with "composition is
% slot-disjoint only" -- true when the engine could NOT read its own pending
% edits, false now that it can. Under exclusion the only thing that could
% invalidate a recorded old is your own earlier same-bracket edit, and RYW now
% reconciles exactly that, so the value/stability distinction is inert here
% rather than surfacing through composition. It earns its keep only in the
% concurrent engine, where a PEER can change a slot between read and commit --
% which is where the guard/validate API lives. Do not re-couple it to
% composition here.
% -----------------------------------------------------------------------

\textbf{Monotonicity is a property of the traversal, not only of the engine.}
\Cref{prop:monotone} holds for a dependency-chained reader. One that hops
without a chain --- re-reading a cached pointer, or comparing two
independently-reached slots --- must supply its own acquire
(\cref{sec:monotone}). The selector's own monotonicity is unconditional; what a
reader's sequence inherits from it is not.

% ===========================================================================
\section{Variations and Alternatives}
\label{sec:variations}

% ---------------------------------------------------------------------------
% THIS SECTION IS THE DEFENSIVE PAYLOAD. A patent claim can cover an obvious
% variant of what is disclosed, so the design space NOT taken must be disclosed
% and dated too. A normal paper cuts this; here it is load-bearing.
%
% RULE: every variant must be ENABLING -- described well enough that a person
% skilled in the art could build it. A variant named but not explained
% discloses nothing.
%
% Do NOT announce the defensive motive in the text. The section stands on its
% academic merit (readers genuinely learn the trade-offs), and the disclosure
% works regardless of stated intent. Announcing it invites hostility and buys
% nothing. If Mathieu wants the motive stated, that is his call -- but the
% default is silence.
%
% The MULTI-WRITER retired variants -- helping, stealing, the per-record install
% latch, invisible writes, the k=1 Bloom filter, the flat budget, the dense-array
% RYW overlay -- belong to P2. They were built, measured and removed, and that
% falsification trail is P2's strongest asset. Do not spend it here.
% ---------------------------------------------------------------------------

The mechanism of \cref{sec:swmechanism} is one point in a design space. This
section maps the space around it: the choices available at each decision, what
each would cost, and why this engine chose as it did. Several of these
alternatives are perfectly workable; the reasons for rejecting them are
quantitative or contextual, not fundamental.

\subsection{The selector}
\label{sec:varselector}

\paragraph{Boolean versus generation counter.} The selector here is a boolean
written once, $0 \to 1$. A generation counter is the obvious alternative: each
proxy records the generation at which its new value becomes current, and
resolution compares the group's counter against it. That buys reusable groups
--- a group could serve many successive transactions without reallocation ---
and it is how clock-based schemes such as RLU decide visibility. It costs the
write-once property, and with it the unconditional monotonicity of
\cref{prop:monotone}: a counter that advances repeatedly makes ``never
new-then-old'' a claim about a range rather than a single transition, and a
reader must then compare rather than test. We chose the boolean because
\cref{prop:monotone} is what lets embedders trade a snapshot for a fixed read
order (\cref{sec:publishorder}), and a boolean makes it free.

\paragraph{One word per group versus packed.} Each group here owns a full word.
Selectors could instead be packed --- many groups' bits in one word, with each
proxy carrying a bit index --- which shrinks the group to nothing but makes the
selector word a contention point shared across unrelated transactions, and makes
each flip a read-modify-write rather than a store. Under exclusion the
read-modify-write is safe; under a concurrent engine it would not be.

\paragraph{Write-once versus reusable.} A group here is used once and reclaimed.
It could be reset and reused, which trades the reclamation cost for a
quiescence problem: the reset must not overtake a reader still resolving a proxy
of the previous use, so the reuse needs exactly the grace period the
reclamation already pays.

\subsection{The tag}
\label{sec:vartag}

\paragraph{Width, and the alignment it implies.} A one-bit tag needs only
two-byte alignment of the record, which any pointer has. The four-bit tag this
engine permits requires sixteen-byte alignment, which must be arranged
deliberately (\cref{sec:alignment}). Wider tags are available at proportionally
stricter alignment: an eight-bit tag would need 256-byte-aligned records, which
is affordable for a small pool of proxies and wasteful for many. The width an
embedder needs is set by its own encoding, not by the engine.

\paragraph{Low bits versus high bits.} This engine takes tag bits from the low
end, where alignment guarantees them. The high end is the alternative: on
current 64-bit architectures the top bits of a virtual address are unused, and a
tag can live there instead, leaving the low bits entirely to the embedder.
The cost is portability --- the number of free high bits is
architecture-dependent and can change --- so a high-bit scheme needs a runtime
probe of the address space to establish its budget, and a fallback for
architectures that offer fewer bits than it wants. An engine could offer both
and let the embedder choose.

\paragraph{Tagging versus a sentinel range.} Tagging is not the only way to make
a proxy recognizable. An embedder could reserve a range of values --- a page of
address space, say, or an unused numeric range --- and treat any value in it as
a proxy handle. This costs no bits of the word, which is the one thing tagging
does cost, and it is a real option for a slot whose bits are all spoken for. It
costs a range check rather than a mask-and-compare on the read path, an
allocator constrained to that range, and a carve-out from the data domain that
the embedder must guarantee is never a legitimate value. For a bitmap
(\cref{sec:nonpointer}) the carve-out is impossible --- every bit pattern is
legitimate data --- which is precisely why the shift-left discipline exists.

\subsection{Proxy placement and allocation}
\label{sec:varproxy}

\paragraph{One block versus many.} Proxies and their group are typically
allocated as a single backing block with one \code{rcu\_head}, so a transaction
costs one allocation and one deferred free. Individually allocated proxies are
the alternative: they cost an allocation and a reclamation each, and they permit
a transaction whose width is not known in advance to grow without moving what it
has already built --- which is the problem \cref{sec:tagnotptr} solves the other
way, by making a move safe.

\paragraph{Referenced versus embedded.} A slot here points at a proxy. A proxy
could instead be embedded in the target object, so the slot points at the object
as usual and the object's header says whether an indirection is in flight ---
which is what an object-header scheme does, and it is the design
\cref{sec:whyweakening} rejects: it reintroduces per-element state, which is the
cost this engine exists to avoid.

\subsection{Settle}
\label{sec:varsettle}

\paragraph{Eager, lazy, or never.} This engine settles eagerly: after the flip,
every slot is rewritten from the tagged proxy back to the direct target. Settle
could be deferred, batched across transactions, or omitted entirely --- the
proxies resolve correctly forever, so a never-settling engine is \emph{correct}.
It is also unbounded: every read of that slot pays an indirection permanently,
the proxies can never be reclaimed, and the structure accumulates a growing
layer of resolved-but-still-present latches. Settle exists to make the cost
transient, which is the whole thesis of \cref{sec:whyweakening}; an engine that
never settled would have per-slot state after all, arrived at by a different
road.

\subsection{Reclamation}
\label{sec:varreclaim}

Proxies are reader-reachable while installed, so their reclamation needs a
scheme that knows when readers have finished. \rcu is one; the others are the
usual suspects and they are usable here. \code{synchronize\_rcu()} instead of
\code{call\_rcu()} makes the writer wait rather than defer, which is simpler and
serializes the writer against every reader in the system. Hazard pointers
\citep{michael2004hazard} would let a proxy be freed as soon as no reader holds
it, at the cost of a per-read store --- which is exactly the read-side charge
this design refuses (\cref{sec:whyweakening}), so it would be a strange choice
here while being a reasonable one elsewhere. Reference counting on the group
would serialize readers on a shared counter.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This subsection matters more than it looks for disclosure purposes: it
% covers the "same mechanism, different reclamation" variants. Keep the
% reasoning attached to each -- "hazard pointers would work but cost a per-
% read store" is enabling; "we use RCU" is not.
% -----------------------------------------------------------------------

\subsection{Non-pointer encodings}
\label{sec:varencoding}

\Cref{sec:nonpointer} stores integers shifted left by one. The alternatives are
to carve a sentinel out of the value domain (impossible for a bitmap, viable for
a counter that never reaches certain values), to keep the payload in a separate
word from the transacted handle (which costs a word and an indirection but frees
the payload entirely), or to widen the slot beyond a machine word and transact
it with a double-width compare-and-swap where the hardware offers one --- which
buys back the spent bit at the cost of portability and of the engine's
word-sized-slot assumption.

% ===========================================================================
\section{Related Work}
\label{sec:related}

\subsection{Publishing is not compare-and-swapping}
\label{sec:relmcas}

Multi-word compare-and-swap is the obvious neighbour and the wrong comparison,
so it is worth separating carefully.

Harris, Fraser and Pratt's practical \mcas \citep{harris2002mcas} makes $k$
words change together under concurrent writers: a descriptor names the records,
threads that meet it \emph{help} drive it, and RDCSS resolves the ordering
hazard between installing a descriptor and reading it. Fraser's thesis and the
journal generalization develop it into a lock-free toolkit
\citep{fraser2004lockfreedom, fraser2007locks}; wait-free constructions follow
\citep{sundell2011waitfreemcas, feldman2015waitfreemcas}; descriptor reuse
\citep{arbelraviv2017descriptors} and lower-overhead \kcas
\citep{guerraoui2020mcas} refine the cost.

The \fliplatch is not one of these. It performs \emph{no compare-and-swap at
all}: under exclusion there is nothing to compare against, no contender to lose
a race to, and no helping to arrange. It solves the narrower problem of atomic
multi-slot \emph{publication} --- making a prepared set of values visible at
once to readers who are not writers --- rather than atomic multi-slot
\emph{update} among competing writers.

What it does inherit from that line is the shape: a descriptor of records plus a
status word whose transition decides the outcome, and resolution routed through
the status rather than through the slot's plain value. That shape is Harris's,
and we claim none of it. The companion paper, which does contend, engages this
work properly.

\subsection{Transactional memory}
\label{sec:relstm}

Software transactional memory \citep{shavit1995stm, herlihy2003dstm} and its
object-based descendants \citep{fraser2004lockfreedom} offer the strictly
stronger contract this engine declines: a transaction is atomic \emph{and}
isolated, so a reader's whole read set is mutually consistent. Opacity
\citep{guerraoui2008opacity} names that reader-side guarantee precisely, and
naming it is what lets us be precise about what we do not provide.

\Cref{sec:fssense} argued that the divergence is not a matter of degree. STM
means the database sense of \emph{transaction}, which bundles isolation into the
word; we mean the journaling sense, which never has. The two are different
contracts, not stronger and weaker versions of one.

Transactional boosting \citep{herlihy2008boosting} is the closest relative on
the composition axis: coarse abstract locks over cheap physical operations,
rather than instrumented reads. It is the pattern a lock-plus-\fliplatch design
converges on, and the companion paper takes it up.

\subsection{Multi-pointer atomicity under RCU}
\label{sec:relrcu}

\rcu \citep{mckenney1998rcu, desnoyers2012urcu, mckenney2023perfbook} gives
readers a free ride and gives writers one atomic pointer store. Extending it to
several pointers at once is an old problem with two established answers, and
\cref{tab:comparison} places both.

\textbf{Existence structures} \citep{mckenney2015issaquah,
mckenney2016beyondissaquah, mckenney2026perfbookcode} give each element a field
naming an \emph{existence group}, and flip the group's state word to change the
membership of every element in it at once. The mechanism is the closest prior
art to this paper: it is a shared word whose single store switches a set's
visibility to \rcu readers, followed by reclamation after a grace period. The
technique has been public since at least January 2015 and was explained at
length in 2016.

\textbf{RLU} \citep{matveev2015rlu} gives each object a header able to name a
newer copy, logs a writer's copies, and makes them visible by advancing a global
clock. \textbf{MV-RLU} \citep{kim2019mvrlu} adds multi-versioning to relieve
RLU's synchronous writeback. Both provide a \emph{coherent snapshot} --- which
is strictly stronger than \cref{prop:monotone}, and must be said plainly: any
comparison of throughput between this engine and RLU crosses guarantee classes.
Against existence structures the classes match and the comparison is clean.

What separates this work from both is where the cost is paid rather than what
the mechanism does at the moment of commit. Existence needs an element to hang a
field on; RLU needs an object to prepend a header to. Both are therefore
permanent additions to the data readers traverse. A \pseudotxn puts its marker
in the slot's own spare bits, which is why it can transact a bare bitmap word
that is not an object at all (\cref{sec:nonpointer}), and why its cost vanishes
when no mutation is in flight.

\subsection{Patented prior art}
\label{sec:relpatents}

The atomic-move-under-\rcu problem has been the subject of a series of patents
naming McKenney and assigned to IBM, and they are prior art whether or not any
remains in force. Four are directly relevant, and their claims bound what this
paper may claim.

\citet{mckenney2008filerename926} (filed 2004) renames a file while permitting
lock-free lookups by linking a \emph{temporary record} into the hash chain and
removing it afterwards --- a transient object, erected for the mutation and gone
after. Its readers, however, \emph{wait}: the claim recites that the temporary
record ``causes the look-ups to wait'' until the rename completes.
\citet{mckenney2010keyupdate851} (filed 2006) moves an element between lists by
making it ``appear simultaneously'' in both under one key and then changing the
key --- appear-in-both, belong-to-one, which is the existence notion in embryo.
\citet{mckenney2018atomicmove907} (filed 2015) atomically updates a
status-indicating entity to transfer an element's validity from an original to a
copy. \citet{mckenney2019atomicmove638} (filed 2016) changes an existence
group's state field ``using a single store operation'' so that, where there are
plural elements, one set is deemed to exist and another deemed not to.

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% Cite the patents as PRIOR ART and stop there. Do not editorialise about
% legal status. All ten in the bib are recorded as lapsed or expired as of
% 2026-07-16, and six of those are fee-lapsed with term still running, which
% under 37 CFR 1.378 can be reinstated -- so "lapsed" is a point-in-time
% observation, not a fact about the future, and it is not this paper's
% business either way. The citation's job is to date the disclosure, not to
% assess enforceability. Saying less here is both safer and more correct.
% -----------------------------------------------------------------------

\subsection{What this paper claims}
\label{sec:relclaims}

Given the above, the claim has to be stated as a conjunction, because each
ingredient separately is old.

The \emph{flip} is not new: a single store switching a set's visibility to \rcu
readers is \citet{mckenney2019atomicmove638} and was public at CppCon in 2016
\citep{mckenney2016beyondissaquah}. The \emph{transient object} is not new: a
per-mutation record, removed afterwards, is \citet{mckenney2008filerename926}
from 2004. \emph{Pointer tagging} is not new; existence structures tag bit~0 of
their own membership field. The \emph{descriptor-and-status shape} is Harris's
\citep{harris2002mcas}.

What we have not found in that prior art is the combination: a marker that is
\textbf{transient} (so the structure's steady state is unchanged),
\textbf{non-blocking for readers} (they resolve through it to old-or-new rather
than waiting on it, unlike \citet{mckenney2008filerename926}), and
\textbf{free of per-element state} (it lives in the slot's spare bits rather
than in a field or header, unlike existence and RLU) --- together with the
consequence that follows: because the marker is per-slot rather than per-object,
a \pseudotxn can span slots whose encodings differ, which forces the per-record
tag (\cref{sec:pertag}) and which we are likewise not aware of in prior work.
% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% "We have not found" and "we are not aware of" are the right register --
% they are honest, and they are what a careful reader will believe. "Novel"
% and "the first" are neither, and the prior-art sweep behind this section
% (10 patents, plus the talks) is exactly what makes the weaker phrasing
% credible rather than evasive. Do not let an editing pass strengthen these
% into claims.
% -----------------------------------------------------------------------

% ===========================================================================
\section{Conclusion}
\label{sec:conclusion}

\rcu gives a writer one atomic act, and the limit is not academic. It is visible
in the interface of a shipped container: Userspace \rcu's list maintains a
backward chain that no reader is allowed to walk, and the missing reverse
iterator is not an oversight but the thing that licenses the plain store to
\code{prev} (\cref{sec:onestore}). Where an update needs two edges to move
together, the remedies have been to copy a larger region, to wait out a grace
period, or to withdraw the operation.

We have described a fourth answer that leaves the bargain intact. A \pseudotxn
is a frozen set of \record{slot}{old}{new} records made visible by one commit
step; it does not add a phase to \rcu's build--publish--reclaim discipline but
widens the middle one, so that a single store moves a set of slots rather than a
single slot (\cref{sec:buildpublish}). The \fliplatch realizes it for a single
writer with no compare-and-swap, no conflict detection and no abort: every proxy
in a group names one shared selector word, and one release store switches the
whole set (\cref{sec:fliplatch}). The marker that makes this possible rides in
spare bits of the slot the reader had to load anyway, so it costs no per-element
state and no additional load, and it is gone once the mutation settles
(\cref{sec:tags}).

The guarantee is deliberately weaker than transactional memory, and we have
tried to be more precise about the weakness than about the strength. The write
set is atomic; a reader is not a transaction of any kind, and a traversal
spanning several
slots may straddle a commit. There is no opacity and no multi-read snapshot.
What a reader gets instead is monotonicity --- never new-then-old
(\cref{sec:monotone}) --- and that middle ground is exploitable rather than
merely tolerable: an embedder that commits in the reverse of its readers'
traversal order gets old-prefix, new-suffix observations and needs no
per-traversal snapshot at all (\cref{sec:publishorder}). The weakening is
targeted. Short of copying the structure outright --- which is the cost that
made the operation expensive to begin with --- isolation is a charge on the
reader, and the reader is what \rcu exists to keep free.

The costs are real and we have stated them where they fall rather than where
they flatter. Exclusion is a precondition of the mechanism, not a caveat on it,
and violating it corrupts silently (\cref{sec:exclusion}). Naming individual
slots means a commit's width grows with the number of edges an operation
re-points, which is the axis on which a design that collapses membership into a
single object will win (\cref{sec:costcomparison}). The tag contract constrains
every value a transacted slot may hold, and an embedder that breaks it does not
get a torn read but a wild pointer (\cref{sec:tagcontract}).

What we claim is a conjunction rather than a component. The marker is transient,
non-blocking for readers, and free of per-element state; each of those exists
separately in prior work, and we have not found them together, nor found the
per-record tag that having them together forces (\cref{sec:related}). We have
described the mechanism, and the variants we did not take
(\cref{sec:variations}), in enough detail to be reimplemented from this paper
alone --- a description that cannot be built from is not a disclosure.

\paragraph{The rest of the series.} This paper is the \sw case, and it is
deliberately the first: the \fliplatch is the model at its smallest, where a
commit is one store and the hard parts are contracts rather than machinery. A
companion paper covers the \mw engine, which attacks the same model with a
multi-word compare-and-swap and pays --- in descriptors, in helping, in an abort
path this paper has no need of --- for the exclusion it drops. The two share the
model of \cref{sec:model} and the status vocabulary, so an embedder that
outgrows exclusion changes which engine it links, not how it reasons. A third
paper takes up the programming model the two engines share: the read policy that
decides which loads must help, read-your-own-writes and the slot-coincidence
hazard it exists to handle, and the declarations an embedder can make to trade
generality for speed. A fourth reports measurements, including where this design
loses and why the loss is the predicted consequence of commit width rather than
a surprise.

% ===========================================================================
\section*{Availability}

% ---------------------------------------------------------------------------
% Pin the COMMIT, never the branch. urcu-txn-dev is a development tip and will
% move; the disclosure is the tree as of this specific commit and nothing else.
% For a defensive publication the dated, specific version IS the deliverable,
% so a reader must be able to fetch exactly what this paper describes years
% from now, without depending on where development happened to go next.
% ---------------------------------------------------------------------------
The engine described here is free software, distributed under
\code{LGPL-2.1-or-later} as part of Userspace \rcu. Every mechanism this paper
describes is present in commit
\code{cefb0414ce36a0443ea425f5fbe183c19fe6db22} (2026-07-17), on branch
\code{urcu-txn-dev} at \url{https://github.com/compudj/userspace-rcu-dev}. The
commit hash, not the branch, identifies the disclosed version: the branch is a
development tip and will move.

Within that tree, the material of this paper is the following headers, all under
\code{include/urcu/}:

\begin{description}\setlength{\itemsep}{2pt}
\item[\code{rcu-txn-sw.h}] The \fliplatch: flip groups, proxies, resolution, and
      the four-step lifecycle of \cref{sec:lifecycle}. Its ordering note is the
      source of \cref{sec:monotone} and \cref{sec:publishorder}, and its
      read-your-own-writes pair (\code{urcu\_txn\_sw\_load} /
      \code{urcu\_txn\_sw\_record\_chain}) is the composition mechanism of
      \cref{sec:swcomposition}.
\item[\code{rcu-txn-sw-list.h}] The bidirectional list of \cref{sec:wrappers},
      whose two chains stay mutually coherent.
\item[\code{rcu-txn-sw-hlist.h}] The hash list.
\item[\code{rcu-txn-sw-bitmap.h}] The transacted bitmap of
      \cref{sec:nonpointer}: a word of 63 logical bits, tag at bit~0.
\item[\code{rcu-txn.h}] The shared model and status vocabulary of
      \cref{sec:model}.
\end{description}

\noindent One further header is named here for orientation rather than as a
contribution of this paper: \code{rcu-mcas.h} is the \mw engine taken up in the
companion paper.

This paper is licensed CC BY 4.0. It is published to place the design, and the
alternatives of \cref{sec:variations}, in the public record as prior art, so
that they remain free for anyone to implement and use.

\bibliographystyle{plainnat}
\bibliography{../common/urcu-txn}

\end{document}
