\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\input{../common/preamble}
\input{../common/macros}

% Draft build: `make draft` drops draft.flag to surface TODO/CITE/NOTE markers.
% Submission builds leave it absent, so the markers vanish.
\IfFileExists{draft.flag}{\draftmodetrue}{}

\title{RCU Pseudo-Transactions and the Sole-Driver MCAS:\\
       A Bounded-Blocking Multi-Writer Engine for\\
       Atomic Multi-Slot Update}

\author{Mathieu Desnoyers\\ EfficiOS Inc.\\ \texttt{mathieu.desnoyers@efficios.com}}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
% ---------------------------------------------------------------------------
% FRAMING RULE FOR THE WHOLE SERIES -- do not soften this in editing.
% This paper refuses to overclaim the PROGRESS class the way the companion
% refused to overclaim the CONSISTENCY model. The engine is bounded-blocking,
% NOT lock-free, and that is a deliberate tail-latency-for-throughput trade.
% Keep "linearizable" scoped to the k-CAS write. NEVER "serializable"/"opacity".
% Overclaiming here is the single most damaging error available to this paper.
% ---------------------------------------------------------------------------
A companion paper described a Read-copy-update (\rcu) \pseudotxn---a frozen set of
\record{slot}{old}{new} records made visible by one commit step---and realized
it for a single writer with a \fliplatch that has no compare-and-swap and
requires writer exclusion. This paper drops the exclusion. The sole-driver
multi-word compare-and-swap (\mcas) engine realizes the same model for concurrent
writers: a transaction is driven only by its owner, and a thread that trips over
an in-flight proxy waits a bounded spin for the owner to settle,
then escalates. There is no helping, no stealing, and---this is the
centerpiece---no RDCSS (restricted double-compare single-swap). A plain value-CAS
install is A-B-A-safe by construction, because the re-plant hazard it prevents
needs a \emph{foreign} driver of one's own records, which sole-driver excludes.

The engine is \emph{bounded-blocking}: it waits on a foreign owner, so it is
not lock-free, and a FIFO-fair mutex fallback makes it starvation-free under a
fair scheduler. We argue this is the right target rather than a shortfall, and
the argument is measured, not hypothesized: a helping design was built and then
removed. Helping forced a per-record install latch that made it
bounded-blocking anyway, so lock-freedom was never actually on the table---and
it lost on throughput.

Per-slot atomicity is not operation atomicity, so a usable concurrent list needs
one thing more: logical deletion, realized as a tombstone that update
transactions agree to treat as freezing a node. It is updater-only
coordination---a would-be attacher validates it and aborts---so a reader gets a
frozen node for free without ever consulting the tombstone, which keeps it off
the reader's path and lets it sit on a cold cache line. This is a known
technique, disclosed here as applied. With it and one guarded read, the same
bidirectional list and hash list become usable under concurrent writers.
\end{abstract}

% ===========================================================================
\section{Introduction}
\label{sec:intro}

Read-copy-update (\rcu) rests on a bargain: readers pay nothing, and in exchange a
writer gets exactly one atomic act, a single pointer store bracketed by building
the new structure where readers cannot reach it and reclaiming the old one after
a grace period~\citep{mckenney1998rcu, desnoyers2012urcu}. A companion paper
gave the \rcu writer a \emph{second} atomic act: a \pseudotxn, a frozen set of
\record{slot}{old}{new} records that a single commit step makes visible at once,
so one store moves a set of slots rather than one. It realized that model for a
single writer with a \fliplatch---a mechanism with no compare-and-swap, no
conflict detection, and no abort, whose price is that it is legal only under
writer mutual exclusion.

This paper drops the exclusion. The question it answers is what the same model
costs when several writers commit concurrently, each without a lock. The answer
is an engine, the sole-driver \mcas (multi-word compare-and-swap), and a claim
about what \emph{kind} of engine it should be.

\paragraph{The thesis.} The companion paper refused to overclaim the model's
\emph{consistency}: it is a \pseudotxn, not software transactional memory, and
we said what isolation it declines and why. This paper refuses to overclaim the
engine's \emph{progress}. It is \emph{bounded-blocking}---it waits, in bounded
steps, on a foreign owner---rather than lock-free, and that is a deliberate
choice of synchronization coarseness: coarse enough to stay simple, portable,
and cache-friendly, fine enough to stay parallel. Bounded-blocking trades a
tail-latency property for throughput knowingly; it is not a failed attempt to
reach lock-freedom. The paper states what it gives up and argues the trade. It
does not apologize for it.

The argument is not a preference stated after the fact. It is a falsification
trail. An earlier version of this engine had threads \emph{help} drive each
other's transactions, in the manner of practical \mcas
\citep{harris2002mcas, barnes1993method}. Helping made two threads co-install a
transaction's records, which admits an install-time A-B-A hazard---a value that
cycles away and back, which a plain compare-and-swap cannot detect---which forced
a per-record \code{FREE}/\code{BUSY}/\code{DONE} install latch to close. That latch
already made the helping engine bounded-blocking---so lock-freedom was never
actually on the table there---and it lost on throughput to the simpler design:
hot-cache-line traffic on the per-record state machine, and helper work that
duplicated the owner's. Removing the helpers removes the co-installer, and with
it the hazard, the latch, and the traffic. What remains is one driver per
transaction and a single residual wait---the coarse, per-transaction settle-wait
of a foreign owner. That is the sole-driver engine, and it wins the measurement
its predecessor lost.

\subsection*{Contributions}

\begin{itemize}
\item \textbf{The sole-driver \mcas engine} (\cref{sec:engine}): a frozen record
      set, a tri-state status word, and descriptor-naming installs, driven by
      the owner alone. Its centerpiece is a \emph{negative} result---the engine
      needs \textbf{no RDCSS} (restricted double-compare single-swap, the
      double-word primitive that normally guards a descriptor install). A plain
      value-CAS install is A-B-A-safe by construction under one driver, because
      the re-plant hazard it resolves requires a foreign driver of one's own
      records, which the regime excludes (\cref{sec:norddcss}).

\item \textbf{The progress argument} (\cref{sec:progress}): deadlock-freedom by
      construction from a sorted install order; \emph{bounded-blocking} defined
      and placed in the progress taxonomy---not obstruction-free, because it
      waits, but starvation-free through a FIFO-fair mutex escalation whose
      budget scales with an operation's cost.

\item \textbf{A measured design-space map} (\cref{sec:variations}): helping,
      abort-others, age-as-resolution, DCAS (double compare-and-swap)/RDCSS, and
      pure lock-freedom---each
      built or considered, each rejected, each with the throughput or
      portability reason that decided it. Disclosing an implemented-then-rejected
      approach is what makes this section carry defensive weight.

\item \textbf{Logical deletion, applied} (\cref{sec:tombstone},
      \cref{sec:structures}): per-slot atomicity is not operation atomicity, so a
      \emph{usable} concurrent structure needs more than a wide commit. We reuse a
      known technique---a logical-deletion tombstone that update transactions
      treat as freezing a node---and disclose the design rationale that suits it
      to this engine: because the freeze is updater-only coordination, a reader
      gets a frozen node for free without consulting the tombstone, so it stays
      off the reader's path. A single guarded read folds the one remaining
      liveness precondition into the commit. We claim no novelty in the
      mechanism; with it, the companion paper's bidirectional list and hash list
      are shown under concurrent writers.
\end{itemize}

\paragraph{Scope.} This paper is about write-set atomicity under concurrency and
what a mutating structure needs to be usable on top of it. The read-side
programming model it forward-references---the read-policy taxonomy, the
general à-la-carte guard facility, and the conflict and aging
\emph{declarations} that tune escalation---is the subject of the next paper in
the series; this one introduces exactly the one guard its own worked example
forces, and no more. The comprehensive comparative evaluation---scaling, tail
latency, and comparison against existence, RLU, and MV-RLU---is later still. What
appears here is only the throughput that justifies each design choice made here.
The engine and the companion \fliplatch share the model and the status
vocabulary, deliberately, so that an embedder that outgrows exclusion changes
which engine it links, not how it reasons about a commit.

% ===========================================================================
\section{Background: what exclusion was buying}
\label{sec:background}

The companion paper's model is assumed and only its vocabulary is recalled here.
A \emph{record} is a triple \record{s}{o}{n}: slot address $s$, the value $o$ it
must hold at the commit point, and the value $n$ it takes afterwards. A
\pseudotxn is a frozen set of such records over pairwise-distinct slots, made
visible by a single commit step. The write set is atomic---every record resolves
to $o$ before the commit and to $n$ after---and no slot and no instant is ever
torn. A reader is not a transaction of any kind: it meets the \pseudotxn one slot
at a time, so a traversal may straddle a commit, but it observes a monotone
sequence and never new-then-old.

The \fliplatch bought all of this cheaply by requiring writer exclusion. Under
exclusion there is nothing to contend for: a commit cannot fail, there is no
install-time compare-and-swap, no conflict detection, and no abort. The whole
mechanism is one shared selector word that a single release store flips from
$0$ to $1$. Exclusion is what makes that legal, and the paper stated it as a
precondition of the mechanism rather than a caveat on it.

Remove exclusion and three things that exclusion was silently handling come due.

\paragraph{Contended installation.} Two writers may now want to change the same
slot at the same time. The engine must let one of them install its intent on the
slot, let the other discover that intent rather than corrupt it, and resolve
which of them a concurrent reader sees---all without a lock. This is the engine's
core job, and \cref{sec:engine} is how it does it: an installed record names a
descriptor, and every reader and every writer resolves the slot through that
descriptor's status word rather than through the slot's plain value.

\paragraph{Per-slot atomicity is not operation atomicity.} A \kcas makes $k$
named slots change together, but a structural operation's correctness can depend
on a slot it does \emph{not} change. Deleting a node and inserting beside it can
touch \emph{different} words and both succeed, and the result is a lost node or a
use-after-free---not because either commit tore, but because the two operations
were never contending for a common slot in the first place. Making a concurrent
list usable therefore needs a discipline layered on the raw engine, and
\cref{sec:tombstone} supplies it: logical deletion.

\paragraph{Install-time A-B-A.} The third hazard is the subtlest, and it is the
one the sole-driver decision turns on: it does not arise from concurrency as
such, but from a specific way of \emph{answering} the first hazard. If threads
help each other install a transaction's records, then a record can be installed
by a thread that is not its owner, and a stale second install can resurrect a
retired descriptor after the slot's value has cycled back to what it names.
\Cref{sec:norddcss} shows that this hazard is helper-induced, so refusing to help
dissolves it rather than merely defending against it---which is why the engine
needs no RDCSS.

% ===========================================================================
\section{The Sole-Driver MCAS Engine}
\label{sec:engine}

\subsection{Records, status, and resolution}
\label{sec:status}

A transaction is a frozen set of records plus one tri-state \emph{status word}:
\begin{center}
\UNDECIDED $\longrightarrow$ \SUCCEEDED\quad(commit)\qquad
\UNDECIDED $\longrightarrow$ \FAILED\quad(abort)
\end{center}
The status word is written once, to a terminal value, and never written back.
Installing a record parks the record's own tagged address in its slot---this is
the descriptor-naming step---and every transition of a slot is such a CAS. A
reader or a writer that loads a slot holding a parked record reads that record's
status: \SUCCEEDED resolves to the record's new value, \UNDECIDED or \FAILED to
its old value. Resolution goes through the status word, so \textbf{a slot's plain
value is never load-bearing}: one transaction's new value is the next
transaction's old value, and only the descriptor's identity disambiguates them.

This is the shape the companion paper inherited from practical \mcas
\citep{harris2002mcas, fraser2004lockfreedom}: a descriptor of records, a status
word whose transition decides the outcome, and resolution routed through the
status rather than through the slot. What is new here is not the shape but what
becomes unnecessary once the single-driver discipline is added to it.

\subsection{Sole-driver: no helping, no stealing, not even reads}
\label{sec:soledriver}

\begin{definition}[Sole-driver discipline]
\label{def:soledriver}
A transaction is installed, decided, and settled by its \emph{owner} alone. No
foreign thread ever plants, drives, evicts, or steals another transaction's
records. A thread that meets a foreign in-flight record resolves it to the
transaction's logical value, or---when it needs the slot itself settled before it
can act on it---waits a bounded spin for the owner to settle it; either way it
proceeds without ever driving the record.
\end{definition}

The discipline is total, and its most easily overlooked edge is the read side. A
reader that lands on an \UNDECIDED foreign record does not help decide it: it
resolves the record to its \emph{logical} value---the old, the value the
transaction takes if it aborts---at once, without waiting on it, and moves on. The
engine's own read accessor states it as an imperative: \emph{never drive} the
foreign transaction. This closes an objection a careful reader will otherwise
raise---``but surely your read path helps''---and it is worth stating outright
because it is what makes the next subsection's argument airtight: there is no
foreign \emph{installer} of a transaction's records, and no foreign \emph{driver}
of any kind, reads included.

A bounded wait is a bias, not a guarantee. It buys the common case, in which the
owner settles promptly; it does not close the window against an owner that has
been preempted mid-transaction. What closes that window is not the wait but the
escalation of \cref{sec:progress}. Keeping the wait and the progress guarantee
separate is deliberate: the wait is an optimization, and the correctness of the
engine does not rest on it.

\subsection{No RDCSS: the centerpiece}
\label{sec:norddcss}

A value-CAS install is the natural way to place a record: compare-and-swap the
slot from the old value the caller saw to the record's tagged address. The
classical objection is an A-B-A: between the caller's read and its install, the
slot could move to some other value and back to the old one, and a naive install
would then succeed against a slot whose meaning had changed underneath it. The
standard defence is RDCSS---a restricted double-compare-single-swap that
conditions the install on a second word, typically the descriptor's own
status---and it is exactly what Harris, Fraser and Pratt introduced to make
practical \mcas safe~\citep{harris2002mcas}.

The sole-driver engine omits RDCSS. It installs with a bare value-CAS, and that
is sufficient. The reason is precise, and it is the paper's strongest claim.

\begin{property}[Install-A-B-A is helper-induced]
\label{prop:abainstall}
The install-time A-B-A that RDCSS prevents requires a \emph{second, stalled
driver of the same transaction}: a thread that installs one of a transaction's
records late, after the slot's value has recurred, resurrecting a descriptor its
owner has already retired. Under \cref{def:soledriver} no such driver exists.
The owner plants its whole prefix before it decides, plants each record exactly
once while the transaction is \UNDECIDED, and no foreign thread ever re-plants,
drives, or settles that record. So a stale second plant cannot occur, and any
recurrence of a slot's plain \emph{value}---a next-pointer cycling
$B \rightarrow X \rightarrow B$ with $B$ a live, never-freed successor, a counter
revisiting a number, an embedder's own reuse---is inert. A losing install is a
plain CAS failure the caller handles by re-reading; it never resurrects anything.
\end{property}

This is applied ownership logic, not a new primitive. The safety of a
disjoint-access-parallel operation that only one thread may drive is the
classical observation of Israeli and Rappoport~\citep{israeli1994disjoint}; the
contribution here is the recognition that a \pseudotxn's install inherits exactly
that safety once helping is removed. The linchpin is that RDCSS's purpose is tied
to helping in the first place: a recent analysis of \kcas frames RDCSS precisely
as the machinery that keeps an install safe \emph{when the operation may be
completed by a helping thread}~\citep{guerraoui2020mcas}. Remove the helping and
the machinery it justified goes with it. We are not aware of prior work that
states the elimination this way---that the sole-driver regime makes RDCSS
unnecessary rather than merely cheaper.

The engine does still require the ordinary things a value-based \mcas requires,
and none of them concerns slot values: tag bit~0 free in every legitimate slot
value (the descriptor-naming tag, \cref{sec:constraints}), pairwise-distinct
slots per transaction, and---the reclamation obligation---that a descriptor
reachable through a slot is freed only after a grace period, because a reader may
still dereference it. This is value-CAS atomicity: a record is validated to hold
its old at the linearization point, not to have been stable throughout. An
embedder that needs stability across an interval layers its own versioning on
top, as with any value-based \mcas. Memory safety holds unconditionally.

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% This is the ENGINE-CLAIMS LEDGER item 2, the centerpiece. Keep it framed as
% (a) a NEGATIVE result -- we removed a mechanism -- and (b) applied ownership
% logic, NOT a new primitive. Israeli-Rappoport for the ownership base;
% Guerraoui 2020 as the linchpin that ties RDCSS to helping. "Apparently
% unpublished" is the register: "we are not aware of", never "novel"/"first".
% Do NOT let an editing pass strengthen the last sentence into a bare claim.
% -----------------------------------------------------------------------

\subsection{Owner-only settle, and where reclamation becomes safe}
\label{sec:settle}

After a transaction is decided, its owner \emph{settles}: it rewrites each record
it planted back to a plain value---the new value on \SUCCEEDED, the old on
\FAILED---so the indirection is gone and later readers resolve directly. Settle
is owner-only and never waits. Because nothing foreign ever planted into those
records, and the owner converts exactly the prefix it planted, one property holds
the instant settle returns: \textbf{no proxy of the transaction names any slot}.

That property is the whole of the reclamation argument, and it is important to
claim exactly it and no more. Reclaiming descriptors after a grace period is not
new---it is the standard \rcu treatment of any reader-reachable object, and it
has been named and benchmarked as an \mcas reclamation baseline over this very
library~\citep{arbelraviv2017descriptors}. What the sole-driver regime adds is
narrower and is a \emph{reachability} statement: because no foreign thread can
plant, drive, or steal a transaction's records, a terminal-and-settled
transaction is a point at which nothing can any longer make its descriptor
reachable, so the descriptor may be handed to \code{call\_rcu()} with no
reference count and no further interlock. The safety of the free is the ordinary
\rcu grace period; what is owner-derived is only the certainty that the free
point has been reached.

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule).
% HIGH overclaim risk (ENGINE-CLAIMS LEDGER item 5). Do NOT claim "reclaim
% descriptors via RCU" -- that is Arbel-Raviv--Brown's named, benchmarked
% baseline over liburcu. Claim ONLY the refcount-free REACHABILITY point:
% terminal-and-settled means nothing foreign can resurrect the records, so it
% is a safe call_rcu point WITHOUT a refcount. The grace period does the
% freeing; sole-driver only certifies the point. Keep this distinction sharp.
% -----------------------------------------------------------------------

\subsection{Constraints}
\label{sec:constraints}

The engine owns tag bit~0 of every transacted slot: an installed record is
recognized by that bit, so every value a slot may legitimately hold must leave
it clear, which two-byte alignment of the pointed-to objects guarantees. The
record set is frozen at commit---membership is fixed before installation
begins---and the slots of one transaction are pairwise distinct. These are the
same shape of contract the companion paper's tag scheme imposed, and they are
discharged by the embedder's encoding, by construction or not at all.

% ===========================================================================
\section{Progress and the Coarseness Thesis}
\label{sec:progress}

\subsection{Deadlock-freedom by construction}
\label{sec:deadlock}

A committer that waits on a foreign owner while holding installs of its own could
deadlock if two committers waited on each other. The engine prevents it not with
detection but with a total order: records install in one global order, sorted by
slot address. A committer therefore holds only \emph{lower} slots while it waits
on a shared one, and the owner it waits on---having itself reached that shared
slot---already passed every lower slot without blocking, so it holds none the
waiter holds. There is no cycle to close, so there is no deadlock.

The sort is paid only when it is needed. A first, optimistic attempt installs
flat, in caller order, and never waits: the moment it meets a foreign record it
bails and fails fast, which cannot deadlock because it never holds-and-waits. Only
when an attempt escalates to the blocking, waiting path does it first sort its
records by address, buying the ordering the wait needs. Age~0 is unsorted and
non-blocking; age~1 and above are sorted and blocking.

\subsection{Bounded-blocking, defined}
\label{sec:boundedblocking}

\begin{definition}[Bounded-blocking]
\label{def:boundedblocking}
An operation is \emph{bounded-blocking} if the only way it waits on another
thread is a wait of bounded duration on that thread to complete a bounded amount
of work, after which the waiter either proceeds or escalates to a mechanism that
guarantees its own progress.
\end{definition}

``Bounded-blocking'' is our descriptive term, not an established progress class,
and it is worth being exact about where it sits. The engine \emph{waits} on a
foreign owner, so it is blocking, and therefore it is not
obstruction-free~\citep{herlihy2003obstruction}: an obstruction-free operation
completes if it runs in isolation for long enough, and this one can be made to
wait even then. That is a real concession and we make it plainly. What the engine
does provide is stronger than deadlock-freedom: through the escalation of
\cref{sec:escalation}, a starved committer eventually acquires a FIFO-fair lane
and completes, so under a fair scheduler the engine is
\emph{starvation-free}~\citep{herlihy2011progress}. Bounded-blocking is thus the
conjunction of a bounded common-case wait and a starvation-free fallback, and it
sits below lock-freedom and above deadlock-freedom in the taxonomy.

\subsection{Adaptive coarseness: the fair-mutex escalation}
\label{sec:escalation}

The optimistic path is bounded-blocking but not by itself starvation-free: a
large transaction can be defeated indefinitely by a stream of small ones that
keep changing a slot between its read and its install. The engine handles this by
escalating a losing handle into a per-domain \emph{fair mutex}---an MCS-style
queue lock~\citep{mellorcrummey1991mcs}---and publishing a flag so that future
transactions in the domain funnel through the same lane. Inside the lane the
escalated transaction contends only with the finite in-flight set, bounded by the
thread count, and commits within a bounded number of retries while holding its
turn. Progress is guaranteed with no quiescence: no \code{synchronize\_rcu()},
no drain.

The escalation is reactive, and its threshold is the interesting part. A handle
takes the lane not because its transaction is large but because it is
\emph{losing}: it has retried past a budget that its own cost earns it. The
budget scales with cost---measured as the transaction's read count plus its
write-set size, the whole bracket rather than only the edges it commits---for a
reason that is itself a small throughput argument. Escalating funnels every
writer in the domain through one serial lane, so the cost of escalating is the
length of the critical section it serializes. A single-edge operation commits
with a bare CAS: serializing it is nearly free, the lane is just orderly queueing,
and escalating early turns wasted CAS collisions into turns. A traversal mutator
is the opposite: funnelling those destroys the $N$-way parallelism of the
optimistic path, which aborts far more often but retries in parallel. On a
three-structure skip-list move, the optimistic path beats the serialized lane by
a factor of roughly $2.6$.\footnote{Measured on the engine's own
\code{bench\_txn\_3skiplist} microbenchmark; see \cref{sec:availability}. The
comprehensive study is deferred to the evaluation paper.} So the more a
transaction costs, the longer it should stay optimistic---the budget encodes
exactly that.

\subsection{Preemption: a design lever, not a claim}
\label{sec:preemption}

A lock-free operation \emph{tolerates} preemption at a standing cost: it is
designed so that a thread stopped at any point obstructs no one. A
bounded-blocking operation instead has a preempted owner as its worst case, the
one the escalation exists to rescue. There is a third option that neither
tolerates nor merely rescues: \emph{avoid} the preemption. The Linux kernel has
gained an \code{rseq} time-slice extension that lets a userspace thread defer its
own preemption for a bounded window~\citep{gleixner2025rseqtse}; applied to the
install-through-settle section of a sole-driver commit, it would let an owner
reach settle before it is taken off-CPU, turning the preempted-owner tail from
\emph{bounded} into \emph{rare}.

We are deliberately careful about what is and is not claimed here.

% ---------------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see README framing rule + SCOPE).
% RSEQ-TSE is NOT in the pinned tree -- verified against rcu-txn.h at the pin:
% no rseq, no time-slice, not even a TODO. So this subsection describes a design
% lever, NOT a measured feature, and the current throughput numbers are the
% PESSIMISTIC case without it. Mathieu's claim boundary (2026-07-18):
%   - claim NOTHING on the mechanism (the mainline rseq-TSE series, Gleixner
%     2025, which the author took part in);
%   - claim NOTHING on the use-pattern "defer preemption of a critical-section
%     holder" -- textbook (schedctl / temporary non-preemption / scheduler-
%     conscious sync); cite the lineage, do not claim it;
%   - DISCLOSE, as defensive prior art, the userspace use for THIS purpose
%     (deferring preemption across the sole-driver install->settle window), and
%     claim only the engine-level OBSERVATION that it thins the tail from
%     bounded to rare -- which is what makes bounded-blocking practically
%     competitive.
% RESOLVED (Mathieu, 2026-07-19): KEEP the subsection (option a) -- explicit
% future/defensive-disclosure, clearly marked unimplemented. The schedctl/TNP
% lineage cites are added and DBLP/patent-verified: schedctl = US Patent
% 5,937,187 (kosche1999preemption) + its unmerged Linux restatement, Aziz LKML
% 2014 (aziz2014preempt); temporary non-preemption = Marsh et al. SOSP
% 1991 (marsh1991firstclass); scheduler activations = Anderson et al. SOSP 1991
% (anderson1991scheduler); lock-holder-preemption avoidance = Dice-Harris
% TRANSACT 2016 (dice2016lockholder).
% ---------------------------------------------------------------------------
The \emph{mechanism} is not ours: it is the mainline kernel series, and the
general pattern of deferring a critical-section holder's preemption so waiters do
not stall is textbook---Solaris's \code{schedctl} preemption
control~\citep{kosche1999preemption} and its later Linux
restatement~\citep{aziz2014preempt}, the temporary non-preemption of
\citet{marsh1991firstclass}, scheduler activations~\citep{anderson1991scheduler},
and lock-holder-preemption avoidance~\citep{dice2016lockholder}. What
we disclose, as prior art rather than claim as invention, is the specific
userspace use---deferring preemption across a bounded-blocking \mcas's publish
window---together with the engine-level observation that it thins the tail. And
we are explicit that this lever is \textbf{not present in the engine measured
here}: the throughput this paper reports is the pessimistic case, \emph{without}
it, so the lever is unclaimed upside rather than a result. Correctness does not
depend on it: it is best-effort, bounded, and not available on every kernel, and
the aging-to-fair-mutex escalation carries the residual either way.

\subsection{The trade, stated plainly}
\label{sec:trade}

The engine gives up lock-freedom's standing guarantee---that no thread's delay
can stall another---and gets, in exchange, a mechanism with fewer synchronization
points, no per-record install state, less hot-cache-line traffic, and a commit
that is a bare CAS per slot in the common case. The concession is a tail: a
preempted owner blocks its waiters until the escalation resolves it. The claim of
this paper is that for the read-dominated structures \rcu serves, and at the
scale where these engines are used, that is the right end of the trade---and the
next section is the evidence, because the losing end was built too.

% ===========================================================================
\section{Variations and Alternatives}
\label{sec:variations}

% ---------------------------------------------------------------------------
% THIS SECTION IS THE DEFENSIVE PAYLOAD and P2's strongest asset: a MEASURED
% design-space map. Each alternative was built (or seriously considered) and
% then rejected for a throughput or portability reason. Disclosing an
% implemented-then-rejected approach enables it as prior art for free.
%
% MEASUREMENT (SCOPE DECISION #1) -- RESOLVED 2026-07-19: the decisive number is
% the help+steal -> sole-driver A/B, +21.7% at 192 cores (to +25.8% under
% contention), from efficios-trie-benchmark txn_vs_existence_skiplist_engine.csv
% (txn-default = help+steal, txn-spin = sole-driver); retire commit a07b67ba.
% abort-others (= stealing) is folded into that bundle; age-as-resolution has no
% isolated run and is rejected on the correctness trap. No \UNVERIFIED remain.
% ---------------------------------------------------------------------------

The mechanism of \cref{sec:engine} is one point in a design space, and unusually
for such a section the neighbouring points are not hypothetical: several were
implemented and measured, then removed. That is what makes the map defensively
load-bearing---an implemented-then-rejected approach is disclosed and dated
here---and it is what makes the coarseness thesis evidence rather than assertion.

\subsection{Helping, and the latch it forced}
\label{sec:varhelping}

The engine once helped. Threads that met an in-flight transaction drove it to
completion, as practical \mcas does~\citep{harris2002mcas, barnes1993method}.
Helping means several threads co-install one transaction's records, and
co-installation is exactly the condition of \cref{prop:abainstall}: a record can
be planted by a non-owner, so a stale late plant can resurrect a retired
descriptor. Closing that needed a per-record install latch---a
\code{FREE}/\code{BUSY}/\code{DONE} state machine on every record, arbitrating
which installer owned it.

The lesson is that this latch already made the helping engine bounded-blocking:
an installer that found a record \code{BUSY} waited on its holder. Helping never
bought lock-freedom; it bought a \emph{more complex} bounded-blocking engine.
And that complexity cost throughput two ways: the per-record state machine was a
hot cache line, contended by every co-installer, and the helper work itself
largely duplicated effort the owner would complete anyway. Dropping the helpers
removes the co-installer, hence the hazard, hence the latch, hence the traffic,
leaving one driver per transaction and a single coarse settle-wait.

The removal is a measured win, and it is what retired the helping path. On a
three-structure skip-list move at 192 cores, the sole-driver engine sustains
$39.9$ against the helping engine's $32.8$ million key-moves per second---$+21.7\%$---and
the margin widens to $+25.8\%$ as the structures shrink and the shared upper
levels grow hotter; a full rebench across the suite found no workload where
helping won.\footnote{\code{bench\_txn\_3skiplist}, width~3, read-your-own-writes
on, jemalloc, best-of-three over one second, on a 2\,$\times$\,96-core EPYC (384
hardware threads). The helping engine is the former default (full help and
steal); the sole-driver engine is the shipped \code{NO\_HELP}\,$+$\,\code{NO\_STEAL}
install. See \cref{sec:availability}.}

\subsection{Abort-others instead of waiting}
\label{sec:varabort}

A committer that meets a foreign transaction in its path could abort that
transaction rather than wait for it. This is obstruction-free contention
management, and it is well charted: DSTM's ``aggressive'' manager aborts any
conflicting transaction on sight~\citep{herlihy2003dstm}, and the priority and
timestamp policies that tame its livelock are Scherer and
Scott's~\citep{scherer2005contention}, whose root is the wound-wait rule of
classical database concurrency control~\citep{rosenkrantz1978woundwait}. This
engine had it, in the form of \emph{stealing}: a higher-ranked transaction evicted
a foreign one and took its slots. Stealing and helping were one subsystem---to
drive a foreign transaction is to help it, to evict one is to steal from it---so
they were removed together, and their combined cost is the measurement of
\cref{sec:varhelping}. The reason it loses is the same either way: aborting a
foreign transaction discards work that was often about to complete, and the retry
it forces costs more than the bounded wait that replaces it.

\subsection{Age-as-resolution}
\label{sec:varage}

A transaction could resolve a contended slot by \emph{version} rather than by
\emph{status}: stamp each transaction with an age, and let a reader or a
committer decide a slot by comparing ages instead of reading a status word. This
is the timestamp-ordering lineage---DSTM's timestamp
manager~\citep{herlihy2003dstm}, again over wound-wait---and it is a different
thing from the version-as-\emph{validity} of TL2 and
LSA~\citep{dice2006tl2, riegel2006lsa}, which we do not adopt and cite only to
distinguish. Age-as-resolution is rejected on a correctness trap, which is
decisive before throughput enters the argument: an age comparison cannot validate
a slot that a transaction reads but does not write, because there is no record to
carry the age, so the scheme cannot express the guarded read of
\cref{sec:structures} without reintroducing a status word anyway---at which point
it is the status engine with an extra field. We did not carry it to a separate
throughput measurement; the trap retired it first.

\subsection{DCAS or RDCSS for the install}
\label{sec:vardcas}

The classical fix for install-time A-B-A is a double-width primitive---DCAS, or
Harris, Fraser and Pratt's RDCSS~\citep{harris2002mcas}. \Cref{sec:norddcss}
showed sole-driver makes it unnecessary, but it was a live alternative before that
argument was found, and its costs are worth recording because they are the reason
it would have been a poor choice even where it is correct: DCAS is not portably
available and RDCSS adds a second conditioning word to every install, which is
hot-cache-line budget spent on a hazard the regime does not actually admit.

\subsection{Pure lock-freedom}
\label{sec:varlockfree}

Finally, one could insist on lock-freedom outright---no bounded wait, no
escalation lane. The measured history is what argues against it: lock-freedom was
unreachable the moment helping forced the per-record install latch
(\cref{sec:varhelping}), because the latch is itself a blocking point. Its one
genuine prize is immunity to preemption, and \cref{sec:preemption} argues that
prize is available more cheaply, for the common case, through preemption
\emph{avoidance} rather than a standing lock-free cost. We do not claim
lock-freedom is unachievable for this problem; we claim that the version of it we
could build lost, and that the reason it is attractive is answered by a cheaper
mechanism.

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}p{0.26\linewidth} p{0.40\linewidth} p{0.24\linewidth}@{}}
\toprule
\textbf{Alternative} & \textbf{Why rejected} & \textbf{Antecedent} \\
\midrule
Helping (co-install) &
  forced a per-record \code{FREE}/\code{BUSY}/\code{DONE} install latch $\Rightarrow$
  bounded-blocking anyway; hot-cache-line traffic; duplicated work. Removing help
  and steal: \textbf{$+21.7\%$} at 192 cores, to $+25.8\%$ under contention &
  \citet{harris2002mcas}; \citet{barnes1993method} \\
Abort-others &
  the same subsystem as helping (steal $=$ evict a foreign txn); removed with it
  (\cref{sec:varhelping}); discards near-complete work &
  DSTM ``aggressive'' \citep{herlihy2003dstm}; \citet{scherer2005contention} $\leftarrow$ wound-wait \citep{rosenkrantz1978woundwait} \\
Age-as-resolution &
  cannot validate a read-not-written slot without a status word---a correctness
  trap, decisive before throughput &
  DSTM timestamp \citep{herlihy2003dstm}; \emph{not} TL2/LSA \citep{dice2006tl2, riegel2006lsa} \\
DCAS / RDCSS install &
  unnecessary under sole-driver (\cref{sec:norddcss}); unportable; hot-cache-line budget &
  RDCSS \citep{harris2002mcas} \\
Pure lock-freedom &
  unreachable once co-install forced the latch; its one prize (preemption immunity) is cheaper via avoidance &
  --- \\
\bottomrule
\end{tabular}
\caption{The measured design space around the sole-driver engine. Each row was
built or seriously considered and then rejected. The decisive help-and-steal
removal ($+21.7\%$, \cref{sec:varhelping}) is the throughput evidence this paper
carries; the remaining rows are rejected on portability or correctness grounds
that hold independently of a number, and the broader comparative study is
deferred. Disclosing each rejected approach is what makes the map defensive prior
art as well as an argument.}
\label{tab:variations}
\end{table}

% ===========================================================================
\section{Logical Deletion}
\label{sec:tombstone}

\Cref{sec:background} named a hazard the engine does not by itself close:
per-slot atomicity is not operation atomicity. A \kcas commits the slots it
names, all together; but a structural operation can be wrong because of a slot it
does \emph{not} name. The discipline that closes it is \emph{logical deletion}, a
known technique; this section states what it does and the design rationale that
fits it to this engine, and \cref{sec:structures} spends it on the worked
structures. It is presented here, at the engine level, because every mutating
structure on the engine needs it---not only the ones shown---and we claim no
novelty in the mechanism itself (\cref{sec:tombstoneclaim}).

\subsection{The hazard, concretely}
\label{sec:hazard}

Consider a doubly-linked list $A \leftrightarrow B \leftrightarrow C$, and two
operations racing: inserting a node after $A$, and deleting $B$. There are two
ways they can interact, and only one of them is closed by the engine's ordinary
slot contention.

The first is closed for free. Removing $B$ must rewrite $B$'s predecessor's
forward pointer---$A$'s \code{next}, the only forward pointer that names the path
through $B$---and an insert after $A$ rewrites $A$'s \code{next} too. So both
operations compare-and-swap the \emph{same} slot, \code{A->next}, with the same
expected old value, and the engine commits at most one of them. The loser fails
its old-value check, aborts, re-reads, and re-descends into a list that is still
coherent. Whichever wins, the outcome is a well-formed list. This race is closed
\emph{structurally}, by the fact that the two operations were contending for a
common slot after all---not by the tombstone.

The second is the real hazard, and it survives a perfect \kcas. An insert after
$A$ rewrites $B$'s \emph{back} pointer to name the new node, but it only
\emph{reads} $B$'s forward pointer---it has no reason to write it. A concurrent
deletion of $B$ writes $B$'s forward pointer and frees $B$. These are
\emph{different} words: the back-edge write and the forward-edge read do not
contend, so both the insert and the delete can commit, and the insert has just
written a back pointer into a node that is being freed. Nothing tore; the two
operations simply never met on a shared slot. Closing this needs two things: a
durable signal that $B$ is being deleted, set where a would-be attacher can check
it (\cref{sec:mark}), and a way to fold that check into the attacher's own commit
so its write is refused if $B$ died (\cref{sec:structures}).

\subsection{The tombstone is updater-only}
\label{sec:mark}

When a node is deleted, the operation that unlinks it also sets a
\emph{logical-deletion tombstone} on the node, as one word of the same multi-word
commit that unlinks it. Where the tombstone lives is a placement choice. It can
be packed into a reserved tag bit of the node's own forward pointer, distinct
from the engine's descriptor tag, which costs no extra word; or it can be a
separate status word, updated as an ordinary transaction record. Both are the
same logical object---a durable per-node ``this node is being deleted'' flag, set
atomically with the unlink and living for the node's grace-period lifetime---and
\cref{sec:tombstoneplacement} returns to why one might prefer either.

Its role is entirely among \emph{updaters}: once the tombstone is set, every
update transaction treats the node as \emph{frozen}, and the commit machinery
enforces the agreement---an update that would touch a frozen node fails to
commit. Two consequences follow, and they are the whole of what the tombstone is
for.

\emph{Deletion is single-updater.} A second delete of the node, or an insert or
delete anchored to it, finds the tombstone set and aborts, terminating with ``no
such entry'' rather than acting on a node that is already leaving. Exactly one
transaction ever logically deletes a node.

\emph{Nothing attaches to a departing node.} An operation that would link a fresh
element onto the node---writing one of the node's edges---must first agree the
node is still live, which it does by validating the tombstone into its own commit
(\cref{sec:guard}). If the tombstone is set, the attaching commit fails, so no
edge is ever installed into a node that is about to be reclaimed. This is the
disjoint-word hazard of \cref{sec:hazard} closed: the attaching write and the
tombstone are different words, but the validate makes the commit fail on the one
while it writes the other.

\subsection{Why readers never touch it, by design}
\label{sec:tombstoneplacement}

A reader consults the tombstone for nothing, and this is deliberate rather than
incidental. The freeze the tombstone announces is \emph{redundant} with the
transactions that enforce it: because every updater agrees not to touch a
tombstoned node, a reader that reached the node before its grace period sees an
object no one will mutate again---a frozen node---\emph{for free}, without loading
the tombstone at all. The updaters' agreement hands the reader exactly what a
per-read tombstone check could have told it, so the check would be redundant.

That the reader is off the tombstone is what makes its placement free, and it is
the series' read-side thesis carried one step past the companion paper. The
companion paper's proxy tag at least rode in the pointer a reader loads---free to
strip, but present in the reader's world. The tombstone need not be present at
all: packed into the forward pointer it is stripped for free on a word the reader
was loading anyway, and moved to a \emph{separate} word it leaves the reader's
forward pointer pristine and can sit on a cache-cold line the reader never
touches. Either way it adds nothing the reader must load, so in the steady state
the read side is byte-for-byte and cycle-for-cycle what it would be with no engine
present.

\subsection{Prior art}
\label{sec:tombstoneclaim}

% ---------------------------------------------------------------------------
% RESOLVED (Mathieu, 2026-07-18): no novelty is claimed for logical deletion.
% Disclose it as an applied known technique; cite Harris (packed marked pointer)
% / EFRvB / SCX finalize (separate-word flag). The one design property worth
% stating -- readers never consult the tombstone, so it can be cold-CL -- is
% DESIGN RATIONALE, not a claim; SCX is the closest prior art (its info field
% readers also never consult). Do NOT reintroduce a "composition" novelty claim,
% the "forwarding to old" framing (readers do not forward through the tombstone
% -- it is updater-only), or the GC-relocation (Brooks/ZGC/Shenandoah/Bronson)
% distinction apparatus, which existed only to defend the retracted forwarding
% claim. The SCOPE's DECISION #4 (claim the tombstone as a composition) is
% RETRACTED -- see the mark-not-proxy memory.
% ---------------------------------------------------------------------------

We claim no novelty in the mechanism; logical deletion is old, in both of its
placements. A marked pointer whose bit logically deletes a node and fails a racing
writer's CAS is Harris~\citep{harris2001lists}; the equivalent as a separate flag
or record interpreted by updaters is the non-blocking-tree discipline of Ellen,
Fatourou, Ruppert and van Breugel~\citep{ellen2010nbbst} and Natarajan and
Mittal~\citep{natarajan2014bst}, and freezing a record set so later updates to it
fail is the \code{SCX} ``finalize'' of Brown, Ellen and
Ruppert~\citep{brown2013scx}. Setting the tombstone as one word of the same
multi-slot commit that unlinks the node---so that logical deletion and physical
unlinking share a single linearization point---is what a \kcas over a tree already
does with such a flag.

One property is worth calling out, as a design rationale rather than a claim: the
tombstone is consulted by updaters alone and never by readers. The nearest prior
art is again \code{SCX}, whose descriptor sits in a field readers never
read---its reader correctness is a linearization argument over plain reads, not a
mark check. What is particular here is the setting: on a read-barrier-free \rcu
base with a sole-driver \mcas, keeping the tombstone entirely off the reader's
path preserves the read side exactly (\cref{sec:tombstoneplacement}) and leaves
its placement free---packed for compactness, or on a cold line for a pristine
reader pointer. We disclose that as the reason the technique fits this engine, and
claim nothing beyond the fit.

% ===========================================================================
\section{The Same Structures, Now Multi-Writer}
\label{sec:structures}

The companion paper showed a bidirectional list and a hash list on the
\fliplatch. The series keeps the same worked structures across papers, so that
what changes from one to the next is exactly the engine capability, not the
example. Here they run under concurrent writers.

\subsection{Same-slot contention, handled by the engine}
\label{sec:contention}

The parts that are purely about two writers wanting the same edge need nothing
beyond \cref{sec:engine}. Adjacent inserts and deletes that re-point a common
\code{next} slot contend on it through descriptor-naming installs; the engine
picks a winner, the losers abort and re-descend, and the aging escalation of
\cref{sec:escalation} carries a persistently unlucky handle to a fair lane. A
composed operation that spans structures folds its edges into one commit through
the same read-your-own-writes discipline the companion paper introduced---which,
being shared by both engines, is not this paper's to claim. There is no
list-specific machinery for contention; there is the engine.

\subsection{The one guarded read}
\label{sec:guard}

The disjoint-word hazard of \cref{sec:hazard} is what the structures must handle
themselves, and it needs exactly one new thing beyond the tombstone. An insert
after a node re-points that node's successor's \emph{back} pointer, but the slot
that serializes the insert against a deletion of the successor is the successor's
\emph{forward} pointer---a slot the insert reads but does not write. Because the
engine installs a transaction's records in slot-address order, it can reach the
back-pointer write before it reaches the shared forward slot, and so drive the
back-edge store against a successor a concurrent deletion is freeing.

The fix is to fold the read of that forward slot into the commit---to
\emph{validate the successor's tombstone}. The insert issues a
\emph{load-validate} on the successor's forward pointer, the slot where a deletion
of the successor sets its tombstone: it reads the value and records a load-only
guard---a record of the form \edit{v}{v}---so that the commit succeeds only if that
slot still resolves to $v$ at the install point. If a concurrent deletion has set
the tombstone in the meantime, the guard's old value no longer matches, the commit
fails, and the insert retries. The guarded read is the read-side companion to the
tombstone: the forward edge is made safe by a \emph{write}-validated slot, the back
edge by a \emph{read}-validated one.

This is deliberately the minimal, functional form of a general facility. The
guard exists here for one reason---to make this paper's own worked example
correct under concurrent writers---and it is presented as exactly that: a single
read folded into a single commit's conflict set. The general read-side model it
is an instance of---a taxonomy of read policies, à-la-carte guards on arbitrary
read-not-written slots, and the declarations that tune them---is the subject of
the next paper, and this paper claims none of that generality. It claims only
that its own list is usable, and shows the one guard that makes it so.

% -----------------------------------------------------------------------
% NOTE (drafting; retained deliberately -- see SCOPE DECISION #5).
% This is the P2/P3 boundary. P2 introduces ONE guard, functionally, as the
% back-edge dual of the mark, to make its OWN flagship example correct under
% MW. It must NOT claim the a-la-carte read-set-validation novelty (that is
% P3-1, whose novelty is the barrier-free-RCU substrate + opt-in-per-slot
% direction). Keep this subsection about "our list needs it", never about
% "here is a general facility". The exact source phrasing is that the
% load-validate of succ->next "exists ONLY to guard the pprev write".
% -----------------------------------------------------------------------

\subsection{Why the hash list transacts its back link, when the single writer
did not}
\label{sec:hlistpprev}

The companion paper's hash list wrote its backward link---each node's
\code{pprev}, a pointer to the slot that names it---with a plain store, on the
grounds that no reader ever loads it and a single updater has no peer to race.
Under concurrent writers that reasoning does not transfer, and the reason is
instructive about the whole difference between the two engines.

For the single writer, ``writer-only'' meant ``private'': a plain store to
\code{pprev} was instantly visible to the sole updater and to no one else, so
there was no window in which it could be seen wrong. Under concurrency
``writer-only'' still means shared: peer mutators read \code{pprev} while
resolving their own edits, and it is lifetime-critical. A plain store cannot be
made safe either eagerly or lazily. Store it before the commit and a lost commit
leaves the store behind, so a live node names a slot inside a node that never
linked. Store it after the commit and there is a window in which \code{pprev} is
not a stale hint but a dangling pointer into another node---and such hints
accumulate on a single slot, unbounded, uncovered by any read-side critical
section because they sit in live nodes indefinitely. So under concurrency
\code{pprev} is transacted: the multi-writer equivalent of the single writer's
``eager'' store is ``atomic with the commit,'' which is the transaction. This is
not a regression; it is the exclusion crutch coming due, in one concrete slot.

\subsection{Reclamation: no re-link before a grace period}
\label{sec:reclaim}

The build--publish--reclaim discipline carries over, with one multi-writer
addition that the single-writer paper explicitly did not have. A deleted node
must not be re-linked---re-inserted, or recycled into a fresh node---before a
grace period, and not only for the usual reason that a reader may still hold it.
The deleting transaction may still hold a parked descriptor in that node's
forward slot, awaiting its owner's settle. An insert builds its new node's links
with plain stores while the node is private; a plain store into a slot that still
holds a parked descriptor is overwritten when that descriptor settles, and the
write is silently lost, leaving a permanently incoherent edge. So a node's
storage must wait out a grace period before reuse exactly as its reclamation
would---the same wait, extended from ``free'' to ``reuse.''

% ===========================================================================
\section{Limitations}
\label{sec:limitations}

The engine's limits divide the same way the companion paper's did: some are
properties of the model, which can be disliked but not violated, and the rest are
contracts on the embedder, which can be violated at a cost.

\paragraph{Per-slot atomicity is not operation atomicity.} A \kcas makes its
named slots change together; it does not make an operation correct with respect
to a slot it does not name. The disjoint-word structural races this admits are
real, and \cref{sec:tombstone,sec:structures} are the discipline that closes them
for the shown structures---the logical-deletion tombstone and the one guarded
read. A different structure with a different coupling owes itself the same kind of
analysis; logical deletion is a general technique, but recognizing which reads a
given operation must guard is the embedder's obligation.

\paragraph{Bounded-blocking, not lock-free.} The engine waits on a foreign owner,
so a preempted owner blocks its waiters until the escalation resolves them
(\cref{sec:boundedblocking}). It is starvation-free under a fair scheduler, not
lock-free, and correctness is independent of the preemption-avoidance lever of
\cref{sec:preemption}, which is not present in the measured engine and rescues
the tail rather than the correctness.

\paragraph{Value-CAS atomicity.} A record is validated to hold its old value at
the linearization point, not to have been stable throughout the transaction. An
operation whose correctness requires detecting an away-and-back change must carry
its own version in the word; snapshot and version semantics are layered on top,
as with any value-based \mcas. Memory safety holds unconditionally; atomicity is
conditional on the embedder's contracts---tag bit~0 free, pairwise-distinct
slots, deferred descriptor reclaim, and the logical-deletion discipline where
operations couple through unwritten slots.

% ===========================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Multi-word compare-and-swap.} The engine's descriptor-and-status shape
is that of practical \mcas~\citep{harris2002mcas, fraser2004lockfreedom,
fraser2007locks}, and its no-RDCSS result is stated against that line: RDCSS is
the machinery~\citep{harris2002mcas} that a recent \kcas
analysis~\citep{guerraoui2020mcas} ties to helping, which is what
\cref{sec:norddcss} removes. Descriptor reuse and lifetime are Arbel-Raviv and
Brown's~\citep{arbelraviv2017descriptors}, whose \rcu reclamation baseline over
this library is the prior art \cref{sec:settle} is careful not to reclaim as its
own.

\paragraph{Blocking that beats lock-free.} That a blocking design can outperform
a lock-free one at scale is not a new observation---it is the lesson of broad
synchronization studies~\citep{david2013everything} and of flat
combining~\citep{hendler2010flatcombining}, which serializes deliberately to win
throughput. We claim the specific engine and its measured falsification trail, not
the general insight; the progress taxonomy we place it in is
standard~\citep{herlihy2003obstruction, herlihy2011progress}, and the fair-lane
fallback is an MCS lock~\citep{mellorcrummey1991mcs}.

\paragraph{Logical deletion.} The tombstone of \cref{sec:tombstone} is applied
prior art, claimed as none of ours. Marked-pointer deletion is
Harris~\citep{harris2001lists}; the flag/record discipline in non-blocking trees
is Ellen et al.~\citep{ellen2010nbbst} and Natarajan and
Mittal~\citep{natarajan2014bst}; freezing a record set against later updates is
\code{SCX}~\citep{brown2013scx}, which is also the nearest prior art for the one
design property we call out---that readers never consult the tombstone
(\cref{sec:tombstoneclaim}).

\paragraph{RCU trees, and reader-visible old versions.} Concurrent updates to an
\rcu search tree with per-node-locked writers are Arbel and Attiya's
Citrus~\citep{arbel2014citrus}; the sole-driver engine replaces those locks with
lock-free \mcas writers over the same \rcu reader discipline. Reader-visible old
versions under concurrent writers are RLU~\citep{matveev2015rlu} and
MV-RLU~\citep{kim2019mvrlu}, which give a coherent snapshot the \pseudotxn model
declines; a throughput comparison against them crosses guarantee classes and is
the evaluation paper's to make carefully. Transactional
boosting~\citep{herlihy2008boosting} is the composition pattern a
lock-plus-engine design converges on.

% ===========================================================================
\section{Conclusion}
\label{sec:conclusion}

The companion paper gave the \rcu writer a second atomic act under exclusion. This
paper removes the exclusion and pays for it in machinery the first had no need
of---a descriptor-naming install, a status word, a bounded wait, an escalation
lane---while refusing to pay for one thing that machinery usually drags in.
Because a transaction is driven only by its owner, the install-time A-B-A that
RDCSS exists to prevent cannot arise, so the engine installs with a bare
value-CAS and needs no RDCSS. That is a negative result, and it is the paper's
centerpiece.

The engine is bounded-blocking, not lock-free, and we have argued that this is
the right target rather than a shortfall. The argument is measured: the
lock-free-adjacent alternative, helping, was built, and it forced a per-record
install latch that made it bounded-blocking anyway and cost it the throughput it
was supposed to buy. Removing the helpers removed the hazard, the latch, and the
traffic at once. What waits now is a single coarse settle-wait, escalating under
pressure into a FIFO-fair lane whose budget scales with an operation's cost, so
that cheap operations serialize early and expensive traversals stay parallel.

Making a concurrent structure usable took one known technique beyond a wide
commit, because per-slot atomicity is not operation atomicity. A deleted node
carries a logical-deletion tombstone, set atomically with its unlink, that update
transactions treat as freezing the node: deletion stays single-updater, and
nothing attaches to a departing node. Because that freeze is updater-only, a
reader gets a frozen node for free without ever consulting the tombstone, so it
stays off the reader's path and can sit on a cold cache line---the read side
unchanged, as the companion paper promised. Its read-side companion is a single
guarded read that folds one liveness precondition into the commit. We claim no
novelty in the technique; with it, the same bidirectional list and hash list the
companion paper used are usable under concurrent writers---the forward edge made
coherent by a shared-slot write, the back edge by a read-validated one, both
coherent at every step.

\paragraph{The rest of the series.} The read-side programming model this paper
forward-references---the read-policy taxonomy, the general guard facility, and the
conflict and aging declarations---is the next paper's, demonstrated on the same
structures. The comprehensive comparative evaluation, against existence, RLU, and
MV-RLU, and across scaling and tail latency, is later still: it does little
defensively, which is why it is last, and it is the strongest academic artifact,
which is why it is worth doing carefully rather than early.

% ===========================================================================
\section*{Availability}
\label{sec:availability}

% ---------------------------------------------------------------------------
% Pin the COMMIT, never the branch (see README). c7f428a8 is the urcu-txn-dev
% tip that carries the read-policy "re-verb from helping to waiting" -- which is
% load-bearing for the sole-driver read story (Sec. 3.2). Verified pushed to
% github-dev/urcu-txn-dev; newer than P1's pin (b5767298).
% ---------------------------------------------------------------------------
The engine described here is free software, distributed under
\code{LGPL-2.1-or-later} as part of Userspace \rcu. Every mechanism this paper
describes is present in commit
\code{c7f428a853ee167954c945bdf86e2337d7711234} (2026-07-18), on branch
\code{urcu-txn-dev} at \url{https://github.com/compudj/userspace-rcu-dev}. The
commit hash, not the branch, identifies the disclosed version: the branch is a
development tip and will move.

Within that tree, the material of this paper is the following headers, all under
\code{include/urcu/}:

\begin{description}\setlength{\itemsep}{2pt}
\item[\code{rcu-mcas.h}] The sole-driver \mcas engine: records, the tri-state
      status word, descriptor-naming installs, the owner-only drive and settle,
      and the no-RDCSS argument in its own header block
      (\cref{sec:engine,sec:norddcss,sec:settle}).
\item[\code{rcu-txn.h}] The transaction front-end: read policy, read-your-own-writes,
      the disjoint and expect-conflict declarations, and the aging-to-fair-mutex
      escalation whose budget scales with operation cost
      (\cref{sec:escalation}).
\item[\code{rcu-txn-list.h}, \code{rcu-txn-hlist.h}] The multi-writer
      bidirectional list and hash list: the logical-deletion tombstone, the one
      load-validate guard, and the transacted back link
      (\cref{sec:tombstone,sec:structures}).
\item[\code{rcu-txn-status.h}] The commit-status enum both front-ends return, so
      an embedder tests a commit the same way whichever engine it drives.
\end{description}

\noindent The fair mutex of \cref{sec:escalation} is \code{fair-mutex.h}, and the
read-your-own-writes lookup filter both engines share is \code{rcu-txn-bloom.h};
the single-writer \fliplatch of the companion paper is \code{rcu-txn-sw.h} and its
structures. The \code{rseq} time-slice-extension lever of \cref{sec:preemption} is
\emph{not} part of this engine and is named there as a design direction, not a
component of the disclosed tree.

This paper is licensed CC BY 4.0. It is published to place the design, and the
alternatives of \cref{sec:variations}, in the public record as prior art, so that
they remain free for anyone to implement and use.

\bibliographystyle{plainnat}
\bibliography{../common/urcu-txn}

\end{document}
